### 2018-10-18 ms.Rmd
add ms.Rmd

---
title: "`phyr`: An R package for model-based phylogenetic analyses"
author: "Daijiang Li, Russell Dinnage, Lucas Nell, Anthony Ives"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::word_document2: 
    toc: no
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
fontsize: 12pt
link-citations: yes
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/methods-in-ecology-and-evolution.csl
bibliography: ref.bib
---

```{r knitr_options, include = FALSE}
if(!require("xfun")) devtools::install_github("yihui/xfun")
if(!require("phyr")) devtools::install_github("daijiang/phyr")
xfun::pkg_attach2(c("knitr", "kableExtra", "dplyr"))
# opts_knit$set(root.dir = normalizePath("../"))
opts_chunk$set(fig.align='center', cache = FALSE, warning = FALSE, 
               message = FALSE, echo = TRUE)
options(digits = 3, width = 88)
```

To render the Rmd file into HTML file, go to Terminal, `cd` change directory to the folder of this manuscript, then type `make`. Then you browser should open the file for you after it is done. You can render to MS Word with `make ms.docx` too if you want (I see no reason for this...)

Note about how to insert citation:

1. Go to [Google Scholar](https://scholar.google.com); click the top left, and select `Settings`; then in the `Bibliography manager` section, check `show links to import citations into BibTeX`, then Save it.
2. Back to the Google Scholar homepage and search the paper you want to cite. E.g. "Generalized linear mixed models for phylogenetic analyses of community structure", under the item, click `Import into BibTeX`; then copy the whole new page (`Cmd + A` then `Cmd + C`).

    ```
    @article{ives2011generalized,
      title={Generalized linear mixed models for phylogenetic analyses of community structure},
      author={Ives, Anthony R and Helmus, Matthew R},
      journal={Ecological Monographs},
      volume={81},
      number={3},
      pages={511--525},
      year={2011},
      publisher={Wiley Online Library}
    }
    ```

3. Open the `ref.bib` in this folder, and paste the citation information there.
4. To cite the paper in the manuscript, copy its key (e.g. `ives2011generalized` here), and inset `[@ives2011generalized]` to where you want to cite it. For multiple papers, use `[@key1; @key2]`.
5. That's it.

Running Title: Model-based phylogenetic analyses

*Summary*

1. Phylogenetic relationships among species are both a challenge and an opportunity for community ecology and comparative biology. Model-based phylogenetic analysis has advantages ... Currently, R functions to conduct these analyses were distributed across different packages and were mostly written in R, which can be very slow with large datasets.
2. To help remedy this situation, we created `phyr`, an R package that collects and updates functions to conduct model-based phylogenetic analyses. By collecting these functions into one package and rewrite core functions in c++, this package can facilitate the use of model-based phylogenetic analyses.
3. This paper presents the main functions of the package and provides an example using simulated data.
4. The `phyr` package provides a unified software environment with improved performance to conduct model-based phylogenetic analyses.

# Introduction

(2-3 paragraphs)

Phylogenetic relationships: importance and challenges

Model-based methods for phylogenetic analysis: advantages

The need of an R package to facilitate model-based phylogenetic analysis

Example of citations: [@r2017], [@ives2011generalized]

# Package architecture

A table with all main functions

Then brief introduction of each topic

## Phylogenetic diversity

## Cor_phy

## Phylogenetic Generalized Linear Mixed Models

### ML

### Bayes

### Other functions

Such as `communityPGLMM.plot.re()`.

# Example analysis

# Conclusion

# References

### 2018-10-21 ms.Rmd
complete section heads
Testing pushing to repository. Just added Tony's new leanpub book to references as a test...
update set up

@@
 
-To render the Rmd file into HTML file, go to Terminal, `cd` change directory to the folder of this manuscript, then type `make`.
-Then you browser should open the file for you after it is done.
-You can render to MS Word with `make ms.docx` too if you want (I see no reason for this...)
-
-Note about how to insert citation:
-
-1.
-Go to [Google Scholar](https://scholar.google.com); click the top left, and select `Settings`; then in the `Bibliography manager` section, check `show links to import citations into BibTeX`, then Save it.
-2.
-Back to the Google Scholar homepage and search the paper you want to cite.
-E.g.
-"Generalized linear mixed models for phylogenetic analyses of community structure", under the item, click `Import into BibTeX`; then copy the whole new page (`Cmd + A` then `Cmd + C`).
-
-    ```
-    @article{ives2011generalized,
-      title={Generalized linear mixed models for phylogenetic analyses of community structure},
-      author={Ives, Anthony R and Helmus, Matthew R},
-      journal={Ecological Monographs},
-      volume={81},
-      number={3},
-      pages={511--525},
-      year={2011},
-      publisher={Wiley Online Library}
-    }
-    ```
-
-3. Open the `ref.bib` in this folder, and paste the citation information there.
-4. To cite the paper in the manuscript, copy its key (e.g. `ives2011generalized` here), and inset `[@ives2011generalized]` to where you want to cite it. For multiple papers, use `[@key1; @key2]`.
-5. That's it.
-
 Running Title: Model-based phylogenetic analyses
@@
 
+Testing citation of Tony's new book [Dinnage]: [@ives2018book] (BTW, n.p. means No Place, and Author is in place of publisher, since the book is self-published)
+
 # Package architecture
@@
 
+# Acknowledgements
+
+# Authors' contributions
+
+# Data Accessibility
+
 # References

### 2018-11-02 ms.Rmd
add main function table [skip ci]

@@
 if(!require("phyr")) devtools::install_github("daijiang/phyr")
-xfun::pkg_attach2(c("knitr", "kableExtra", "dplyr"))
+xfun::pkg_attach2(c("knitr", "kableExtra", "kableExtra", "dplyr"))
 # opts_knit$set(root.dir = normalizePath("../"))
@@
                message = FALSE, echo = TRUE)
-options(digits = 3, width = 88)
+options(digits = 3, width = 88, knitr.kable.NA = '')
 ```
@@
 
-A table with all main functions
+Main functions provided by `phyr` can be found in Table \@ref(tab:mainFunc).
+
+(ref:maunFuncCaption) Main functions in `phyr`.
+
+```{r mainFunc, echo=FALSE}
+read.csv("phyr_func.csv", check.names = FALSE) %>% 
+  knitr::kable(caption = "(ref:maunFuncCaption)", booktab = T) %>%
+  kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
+                latex_options = c("striped")) %>% 
+  kableExtra::column_spec(1, width = "20%")
+```
 

### 2018-11-02 ms.Rmd
update pd

--- 2018-10-21 ms.Rmd
@@
     number_sections: no
-    theme: readable
+    theme: sandstone
     toc: yes
+    css: style.css
   bookdown::word_document2: 
@@
 if(!require("phyr")) devtools::install_github("daijiang/phyr")
-xfun::pkg_attach2(c("knitr", "kableExtra", "dplyr"))
+xfun::pkg_attach2(c("phyr", "knitr", "kableExtra", "kableExtra", "tidyverse"))
 # opts_knit$set(root.dir = normalizePath("../"))
@@
                message = FALSE, echo = TRUE)
-options(digits = 3, width = 88)
+options(digits = 3, width = 88, knitr.kable.NA = '')
+
+comm = phyr::comm_a
+set.seed(1234)
+comm = comm[sample(nrow(comm), 5), sample(ncol(comm), 5)]
+comm$site = row.names(comm)
+dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
+  left_join(phyr::envi, by = "site") %>% 
+  left_join(phyr::traits, by = "sp")
+dat$pa = as.numeric(dat$freq > 0)
+dat$freq2 = 20 - dat$freq
+pglmm_example = phyr::communityPGLMM(freq ~ 1 + shade + (1 | sp__) + (1 | site) + (1 | sp__@site), 
+                     data = dat, tree = phyr::phylotree)
 ```
@@
 
-Example of citations: [@r2017], [@ives2011generalized]
-
-Testing citation of Tony's new book [Dinnage]: [@ives2018book] (BTW, n.p. means No Place, and Author is in place of publisher, since the book is self-published)
+Example of citations: [@r2017], [@ives2011generalized; @ives2018book]
 
@@
 
-A table with all main functions
+Main functions in `phyr` can be categorized into three groups (Table \@ref(tab:mainFunc)): A) phylogenetic diversity (alpha and beta), B) species' characteristics (e.g.
+correlations among traits, relationships between traits and environmental variables), and C) community assembly.
+
+(ref:maunFuncCaption) Main functions in `phyr`.
+
+```{r mainFunc, echo=FALSE}
+read.csv("phyr_func.csv", check.names = FALSE) %>% 
+  knitr::kable(caption = "(ref:maunFuncCaption)", booktab = T) %>%
+  kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
+                latex_options = c("striped")) %>% 
+  kableExtra::column_spec(2, width = "25%")
+```
 
@@
 
-## Cor_phy
+The set of functions to calculate phylogenetic alpha (`psv`, `pse`, `psr`, `psc`, and `psd`) and beta (`pcd`) diversity was imported from the R package `picante` [@kembel2010picante].
+Their theoretic bases were provided by @helmus2007phylogenetic (for alpha diversity) and @ives2010phylogenetic (for pairwise beta diversity).
+We improved the speed of these functions by rewriting key parts of the codes with c++.
+We added one argument `prune.tree` to the set of phylogenetic alpha diversity functions to allow users to decide whether to prune the phylogeny before converting it into a variance-covariance matrix; its default value is `FALSE`, which means that the phylogeny will first been converted into a variance-covariance matrix; the matrix was then subsetted to the keep only the species interested.
+These two methods of generating the variance-covariance matrix may have different results in some situations.
+We set the default to `FALSE` because this may give a more accurate matrix.
 
-## Phylogenetic Generalized Linear Mixed Models
+## Species' characteristics
+
+`cor_phy`
+
+`binaryPGLMM`
+
+## Community assembly
+
+`communityPGLMM` or `pglmm`
 
@@
 
-### Other functions
+### Other help functions
 
-Such as `communityPGLMM.plot.re()`.
+Such as `communityPGLMM.plot.re()`. We can use it to show the design matrices' structures of random terms in `communityPGLMM()`. An example can be shown in Figure \@ref(fig:reStructure).
+
+(ref:reStructureCaption) A figure to show the structures of design matrices of random terms in `communityPGLMM`.
+
+```{r reStructure, fig.width=6, fig.height=6, fig.cap="(ref:reStructureCaption)"}
+communityPGLMM.plot.re(x = pglmm_example, show.image = TRUE, show.sim.image = FALSE)
+```
 

### 2018-11-02 ms.Rmd
update pglmm [skip ci]

@@
 link-citations: yes
+always_allow_html: yes
 csl: https://raw.githubusercontent.com/citation-style-language/styles/master/methods-in-ecology-and-evolution.csl
@@
 
-Then brief introduction of each topic
-
 ## Phylogenetic diversity
@@
 
-`communityPGLMM` or `pglmm`
+The original code for `communityPGLMM` were imported from the R package `pez` [@pearse2015pez].
+Its theorectical background was presented in @ives2011generalized.
+We improved both the user interface and the speed of `communityPGLMM` here.
+Function `pez::communityPGLMM` requires user-prepared random terms while `phyr::communityPGLMM` makes this step much easier by adapting formula syntax for mixed models provided by the R package `lme4` [@bates2015lme4].
 

### 2018-11-03 ms.Rmd
update pglmm [skip ci]

@@
 Its theorectical background was presented in @ives2011generalized.
-We improved both the user interface and the speed of `communityPGLMM` here.
-Function `pez::communityPGLMM` requires user-prepared random terms while `phyr::communityPGLMM` makes this step much easier by adapting formula syntax for mixed models provided by the R package `lme4` [@bates2015lme4].
+We made several important updates here.
+
+- First, function `pez::communityPGLMM` requires user-prepared random terms while `phyr::communityPGLMM` makes this step much easier by adapting formula syntax for mixed models provided by the R package `lme4` [@bates2015lme4].
+To specify a random term with design matrix derived from a phylogenetic variance-covariance matrix, we can append two underscores (`__`) to the group variable.
+For example, `(1 | sp__)` will create *two* random terms: one with and one without (i.e.
+identity matrix) phylogenetic design matrix.
+To test whether closely related species are more or less likely to co-occur in the same site, we can use `(sp__@site)` along with the argument `repulsion = TRUE` (overdispersion) or `repulsion = FALSE` (underdispersion).
+- Second, we extended `communityPGLMM` to work with Binomial and Poisson distributed data while `pez::communityPGLMM` only allows Binary and Gaussian data.
+- Third, we provided both Maximum likelihood and Bayesian approaches to fit PGLMMs.
+Users can choose the approach via `bayes = TRUE` or `bayes = FALSE` (default).
+The Bayesian PGLMMs were fitted with integrated nested Laplace approximations [INLA, @rue2009approximate].
+- Fourth, we improved the speed of `communityPGLMM` by rewriting key functions in c++.
 
@@
 
+### Speed comparisons
+
+A table to compare the speed of functions in `phyr` with their original versions.
+
 # Example analysis

### 2018-11-07 ms.Rmd
update ms [skip ci]

--- 2018-11-02 ms.Rmd
@@
 ```{r knitr_options, include = FALSE}
-if(!require("xfun")) devtools::install_github("yihui/xfun")
+if(!require("xfun")) install.packages("xfun")
 if(!require("phyr")) devtools::install_github("daijiang/phyr")
@@
 
-`cor_phy`
+Two functions (`cor_phy` and `binaryPGLMM`) were included in this group. Both of them were included in the R package `ape` [@ape2018] orginally.
 
-`binaryPGLMM`
+`cor_phy`: ... (Lucas?)
+
+`binaryPGLMM`: only 1 value per species, right?
 
@@
 Its theorectical background was presented in @ives2011generalized.
-We improved both the user interface and the speed of `communityPGLMM` here.
-Function `pez::communityPGLMM` requires user-prepared random terms while `phyr::communityPGLMM` makes this step much easier by adapting formula syntax for mixed models provided by the R package `lme4` [@bates2015lme4].
+We made several important updates here.
+First, function `pez::communityPGLMM` requires user-prepared random terms while `phyr::communityPGLMM` makes this step much easier by adapting formula syntax for mixed models provided by the R package `lme4` [@bates2015lme4].
+To specify a random term with design matrix derived from a phylogenetic variance-covariance matrix, we can append two underscores (`__`) to the group variable.
+For example, `(1 | sp__)` will create *two* random terms: one with and one without (i.e.
+identity matrix) phylogenetic design matrix.
+To test whether closely related species are more or less likely to co-occur in the same site, we can use `(sp__@site)` along with the argument `repulsion = TRUE` (overdispersion) or `repulsion = FALSE` (underdispersion; default).
+Second, we extended `communityPGLMM` to work with Binomial and Poisson distributed data while `pez::communityPGLMM` only allows Binary and Gaussian data.
+Third, we provided both Maximum likelihood and Bayesian approaches to fit PGLMMs.
+Users can choose the approach via `bayes = TRUE` or `bayes = FALSE` (default).
+The Bayesian PGLMMs were fitted with integrated nested Laplace approximations (INLA) via the R package `r-inla` [@rue2009approximate].
+Fourth, we improved the speed of `communityPGLMM` by rewriting key functions in c++.
 
-### ML
+The function `communityPGLMM` is modularized (prepare data, design matrix, optimization, etc.). (Too much details??)
 
-### Bayes
-
-### Other help functions
-
-Such as `communityPGLMM.plot.re()`. We can use it to show the design matrices' structures of random terms in `communityPGLMM()`. An example can be shown in Figure \@ref(fig:reStructure).
+The class of the model object returned is `communityPGLMM`, which has multiple methods (e.g., `print`, `summary`) associated with.
+Use `methods(class = 'communityPGLMM')` to see the full list.
+To visualize the design matrices' structures of random terms in the model object, we can use `communityPGLMM.plot.re()`.
+An example can be shown in Figure \@ref(fig:reStructure).
+It is also optional to simulate and plot data based on the design matrices with the argument `show.sim.image = TRUE`.
 
@@
 
+### Speed comparisons
+
+A table to compare the speed of functions in `phyr` with their original versions.
+
 # Example analysis
+
+Simulation of data; then apply the main functions.
+
+# Limitations
+
+`communityPGLMM` requires columns `site` and `sp`.
 

### 2018-11-16 ms.Rmd
Added small blurb about cor_phylo [skip ci]

@@
 
-Two functions (`cor_phy` and `binaryPGLMM`) were included in this group. Both of them were included in the R package `ape` [@ape2018] orginally.
+Two functions (`cor_phylo` and `binaryPGLMM`) were included in this group. Both of them were included in the R package `ape` [@ape2018] orginally.
 
-`cor_phy`: ... (Lucas?)
+The function `cor_phylo` computes Pearson correlation coefficients for multiple continuous
+traits that may have phylogenetic signal, allowing users to also specify measurement
+error.
+The underlying methods were originally developed by @zheng2009new.
+We improved speed by writing many of the internal functions in c++,
+made the input format more flexible,
+fixed a mistake that meant measurement error was not included, and
+added two new features: parametric bootstrapping and the ability to use optimizers
+from the `nlopt` library [@johnson2014nlopt; @powell2009bobyqa; @nelder1965simplex; @rowan1990functional].
 

### 2019-06-24 ms.Rmd
update [skip ci]

@@
 comm = phyr::comm_a
-set.seed(1234)
-comm = comm[sample(nrow(comm), 5), sample(ncol(comm), 5)]
 comm$site = row.names(comm)
+# add site environmental data and species traits
 dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
@@
 pglmm_example = phyr::communityPGLMM(freq ~ 1 + shade + (1 | sp__) + (1 | site) + (1 | sp__@site), 
-                     data = dat, tree = phyr::phylotree)
+                     data = dat, cov_ranef = list(sp = phyr::phylotree))
 ```
@@
 3. This paper presents the main functions of the package and provides an example using simulated data.
-4. The `phyr` package provides a unified software environment with improved performance to conduct model-based phylogenetic analyses.
+4. The `phyr` package provides a unified software environment with improved performance to conduct model-based phylogenetic analyses in Ecology.
 
@@
 
-(ref:maunFuncCaption) Main functions in `phyr`.
+(ref:maunFuncCaption) List of main functions in `phyr` package.
 
@@
 
-## Species' characteristics
+## Species' traits
 
@@
 
-The function `cor_phylo` computes Pearson correlation coefficients for multiple continuous
-traits that may have phylogenetic signal, allowing users to also specify measurement
-error.
+The function `cor_phylo` computes Pearson correlation coefficients for multiple continuous traits that may have phylogenetic signal, allowing users to also specify measurement error.
 The underlying methods were originally developed by @zheng2009new.
-We improved speed by writing many of the internal functions in c++,
-made the input format more flexible,
-fixed a mistake that meant measurement error was not included, and
-added two new features: parametric bootstrapping and the ability to use optimizers
-from the `nlopt` library [@johnson2014nlopt; @powell2009bobyqa; @nelder1965simplex; @rowan1990functional].
+We improved speed by writing many of the internal functions in c++, made the input format more flexible, fixed a mistake that meant measurement error was not included, and added two new features: parametric bootstrapping and the ability to use optimizers from the `nlopt` library [@johnson2014nlopt; @powell2009bobyqa; @nelder1965simplex; @rowan1990functional].
 
@@
 
-### Speed comparisons
-
-A table to compare the speed of functions in `phyr` with their original versions.
-
 # Example analysis
@@
 
-# Limitations
+## Phylogenetic diversity
 
-`communityPGLMM` requires columns `site` and `sp`.
+```{r}
+tree = phyr::phylotree
+phyr::psv(comm_b, tree) # comm_b is a site by species matrix
+phyr::psd(comm_b, tree)
+x = phyr::pcd(comm_b, tree, verbose = FALSE)
+```
 
-# Conclusion
+## Species' traits
+
+## Community assembly
+
+```{r}
+pglmm_example
+```
+
+# Benchmark
+
+A table to compare the speed of functions in `phyr` with their original versions.
+
+```{r, echo=FALSE}
+# data prep for pez::communityPGLMM, not necessary for phyr::communityPGLMM
+dat = arrange(dat, site, sp)
+nspp = n_distinct(dat$sp)
+nsite = n_distinct(dat$site)
+
+dat$site = as.factor(dat$site)
+dat$sp = as.factor(dat$sp)
+
+tree = ape::drop.tip(phylotree, setdiff(phylotree$tip.label, unique(dat$sp)))
+Vphy <- ape::vcv(tree)
+Vphy <- Vphy/max(Vphy)
+Vphy <- Vphy/exp(determinant(Vphy)$modulus[1]/nspp)
+Vphy = Vphy[levels(dat$sp), levels(dat$sp)]
+
+# prepare random effects
+re.site <- list(1, site = dat$site, covar = diag(nsite))
+re.sp <- list(1, sp = dat$sp, covar = diag(nspp))
+re.sp.phy <- list(1, sp = dat$sp, covar = Vphy)
+# sp is nested within site
+re.nested.phy <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
+re.nested.rep <- list(1, sp = dat$sp, covar = solve(Vphy), site = dat$site) # equal to sp__@site
+# can be named 
+re = list(re.sp = re.sp, re.sp.phy = re.sp.phy, re.nested.phy = re.nested.phy, re.site = re.site)
+```
+
+```{r eval=T}
+microbenchmark::microbenchmark(phyr::pcd(comm = comm_a, tree = phylotree, reps = 1000, verbose = F),
+                               picante::pcd(comm = comm_a, tree = phylotree, reps = 1000),
+                               times = 30)
+rt = ape::rtree(n = 1000)
+microbenchmark::microbenchmark(phyr::vcv2(rt, corr = T),
+                               ape::vcv(rt, corr = T))
+
+microbenchmark::microbenchmark(phyr::psv(comm_a, phylotree, compute.var = T),
+                               picante::psv(comm_a, phylotree, compute.var = T))
+
+microbenchmark::microbenchmark(
+  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
+                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = T, 
+                       optimizer = "bobyqa"),
+    phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
+                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = T, 
+                       optimizer = "Nelder-Mead"),
+  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
+                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = F, 
+                       optimizer = "Nelder-Mead"),
+  pez::communityPGLMM(pa ~ 1 + shade, data = dat, family = "binomial", sp = dat$sp, 
+                      site = dat$site, random.effects = re, REML = F),
+  times = 10
+)
+```
+
+# Implications
+
+# Resources
+
+This package is available at Github.
+
+# Future directions
+
+In the future, we intend to ...
 

### 2019-06-25 ms.Rmd
community ecology focus? [skip ci]

@@
 ---
-title: "`phyr`: An R package for model-based phylogenetic analyses"
+title: "`phyr`: An R package for model-based phylogenetic analyses in community ecology"
 author: "Daijiang Li, Russell Dinnage, Lucas Nell, Anthony Ives"
@@
 
+An ecological community contains multiple species occupying the same geographical area, with all species evolved from the same common ancestor.
+Therefore, for almost every question about ecological community, we need to consider the evolutionary history of species within that community.
+For some questions, it is statistically necessary to do so [citations]; for other questions, evolutionary relationships among species can provide additional insights about the ecological community [citations].
+Consequently, studies in community ecology that involve phylogenies increased exponentially over the past decades [citations].
+Most of these studies fall into one or more of the following groups: 1) phylogenetic diversity of the community, 2) comparative studies of species' traits and their relationships with each other as well as environmental conditions, and 3) community assembly patterns and processes.
+
 Phylogenetic relationships: importance and challenges
 
-Model-based methods for phylogenetic analysis: advantages
+Model-based methods for phylogenetic analysis: advantages and challenges
 
@@
 
-Main functions in `phyr` can be categorized into three groups (Table \@ref(tab:mainFunc)): A) phylogenetic diversity (alpha and beta), B) species' characteristics (e.g.
+`phyr` is written in R and C++.
+Most of the functions in `phyr` have been included in other R packages.
+However, the original versions were all written in pure R and thus are relatively slow.
+Furthermore, methods behind these functional were developed in the same theoretic framework.
+It makes more sense to combine them together: thus the `phyr` package.
+Main functions in `phyr` received new major updates and developments and can be categorized into three groups (Table \@ref(tab:mainFunc)): A) phylogenetic diversity (alpha and beta), B) species' traits (e.g.
 correlations among traits, relationships between traits and environmental variables), and C) community assembly.

### 2019-06-26 ms.Rmd
update [skip ci]

@@
 
-Two functions (`cor_phylo` and `binaryPGLMM`) were included in this group. Both of them were included in the R package `ape` [@ape2018] orginally.
+Two functions (`cor_phylo` and `binaryPGLMM`) were included in this group. Both of them were included in the R package `ape` [@ape2018] originally.
 
@@
 
-```{r eval=T}
+```{r eval=F}
 microbenchmark::microbenchmark(phyr::pcd(comm = comm_a, tree = phylotree, reps = 1000, verbose = F),

### 2019-09-19 ms.Rmd
update ms

--- 2019-06-24 ms.Rmd
@@
 ---
-title: "`phyr`: An R package for model-based phylogenetic analyses"
+title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
 author: "Daijiang Li, Russell Dinnage, Lucas Nell, Anthony Ives"
@@
     css: style.css
+    keep_md: yes
   bookdown::word_document2: 
@@
 options(digits = 3, width = 88, knitr.kable.NA = '')
-
-comm = phyr::comm_a
-comm$site = row.names(comm)
-# add site environmental data and species traits
-dat = tidyr::gather(comm, key = "sp", value = "freq", -site) %>% 
-  left_join(phyr::envi, by = "site") %>% 
-  left_join(phyr::traits, by = "sp")
-dat$pa = as.numeric(dat$freq > 0)
-dat$freq2 = 20 - dat$freq
-pglmm_example = phyr::communityPGLMM(freq ~ 1 + shade + (1 | sp__) + (1 | site) + (1 | sp__@site), 
-                     data = dat, cov_ranef = list(sp = phyr::phylotree))
+# traits, phy, dat
+source(file = "simulation_pglmm.R", echo = FALSE, print.eval = FALSE)
 ```
@@
 
+_Keywords_: phylogenetic diversity, phylogenetic generalized linear mixed models, functional trait, trait correlation, model-based methods
+
 # Introduction
 
-(2-3 paragraphs)
+Ecological communities are collections of species that occur within the same geographical area.
+Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community.
+These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep].
+For example, an insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common.
+Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species.
+For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake.
+The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition.
 
-Phylogenetic relationships: importance and challenges
+Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity.
+The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species.
+Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured.
+For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured.
+Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain.
 
-Model-based methods for phylogenetic analysis: advantages
+How species respond to environmental factors, and how they respond to each other, depend on their traits.
+Therefore, relationships among functional traits can provide insights about the evolutionary history that has shaped species trait so they can occupy the same sites.
+For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH.
+Is the positive correlation between these two traits caused by correlated selective forces?
+A challenge to answering this question is that phylogenetic correlations between trait values of closely related species can give the appearance of positive correlations between traits: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related.
+Therefore, it is important to account for species’ evolutionary history when studying correlations among functional traits.
 
-The need of an R package to facilitate model-based phylogenetic analysis
+Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities.
+The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities.
+They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities.
+These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow.
+The R package `phyr` is designed to overcome many of the costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance.
+Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`.
+We then compare them to methods and programs that are currently available.
+Finally, we give applications of `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.
 
-Example of citations: [@r2017], [@ives2011generalized; @ives2018book]
+# Overview of 'phyr'
 
-# Package architecture
-
-Main functions in `phyr` can be categorized into three groups (Table \@ref(tab:mainFunc)): A) phylogenetic diversity (alpha and beta), B) species' characteristics (e.g.
-correlations among traits, relationships between traits and environmental variables), and C) community assembly.
+`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and phylogenetic diversity metrics (e.g.
+`psv()`, `pse()`).
+The workhorse functions of all groups are written in C++ to increase computational speed.
+Here, we will focus on the first two groups of functions because they are more complicated than phylogenetic diversity metrics.
 
@@
 
-## Phylogenetic diversity
+## _`pglmm()`_
 
-The set of functions to calculate phylogenetic alpha (`psv`, `pse`, `psr`, `psc`, and `psd`) and beta (`pcd`) diversity was imported from the R package `picante` [@kembel2010picante].
-Their theoretic bases were provided by @helmus2007phylogenetic (for alpha diversity) and @ives2010phylogenetic (for pairwise beta diversity).
-We improved the speed of these functions by rewriting key parts of the codes with c++.
-We added one argument `prune.tree` to the set of phylogenetic alpha diversity functions to allow users to decide whether to prune the phylogeny before converting it into a variance-covariance matrix; its default value is `FALSE`, which means that the phylogeny will first been converted into a variance-covariance matrix; the matrix was then subsetted to the keep only the species interested.
-These two methods of generating the variance-covariance matrix may have different results in some situations.
-We set the default to `FALSE` because this may give a more accurate matrix.
+`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species.
+The syntax for `pglmm()` resembles that used in the R package `lme4`, and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`.
+`pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be a phylogenetic covariance matrix or any other covariance matrix that the user defines (e.g., spatial or temporal autocorrelation matrix).
+`pglmm()` can also fit models with “nested” covariance structures (e.g.
+species phylogenetic covariance matrix nested within site covariance matrix).
+Finally, `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions.
 
-## Species' traits
+A general example of the syntax for `pglmm()` is
 
-Two functions (`cor_phylo` and `binaryPGLMM`) were included in this group. Both of them were included in the R package `ape` [@ape2018] orginally.
-
-The function `cor_phylo` computes Pearson correlation coefficients for multiple continuous traits that may have phylogenetic signal, allowing users to also specify measurement error.
-The underlying methods were originally developed by @zheng2009new.
-We improved speed by writing many of the internal functions in c++, made the input format more flexible, fixed a mistake that meant measurement error was not included, and added two new features: parametric bootstrapping and the ability to use optimizers from the `nlopt` library [@johnson2014nlopt; @powell2009bobyqa; @nelder1965simplex; @rowan1990functional].
-
-`binaryPGLMM`: only 1 value per species, right?
-
-## Community assembly
-
-The original code for `communityPGLMM` were imported from the R package `pez` [@pearse2015pez].
-Its theorectical background was presented in @ives2011generalized.
-We made several important updates here.
-First, function `pez::communityPGLMM` requires user-prepared random terms while `phyr::communityPGLMM` makes this step much easier by adapting formula syntax for mixed models provided by the R package `lme4` [@bates2015lme4].
-To specify a random term with design matrix derived from a phylogenetic variance-covariance matrix, we can append two underscores (`__`) to the group variable.
-For example, `(1 | sp__)` will create *two* random terms: one with and one without (i.e.
-identity matrix) phylogenetic design matrix.
-To test whether closely related species are more or less likely to co-occur in the same site, we can use `(sp__@site)` along with the argument `repulsion = TRUE` (overdispersion) or `repulsion = FALSE` (underdispersion; default).
-Second, we extended `communityPGLMM` to work with Binomial and Poisson distributed data while `pez::communityPGLMM` only allows Binary and Gaussian data.
-Third, we provided both Maximum likelihood and Bayesian approaches to fit PGLMMs.
-Users can choose the approach via `bayes = TRUE` or `bayes = FALSE` (default).
-The Bayesian PGLMMs were fitted with integrated nested Laplace approximations (INLA) via the R package `r-inla` [@rue2009approximate].
-Fourth, we improved the speed of `communityPGLMM` by rewriting key functions in c++.
-
-The function `communityPGLMM` is modularized (prepare data, design matrix, optimization, etc.). (Too much details??)
-
-The class of the model object returned is `communityPGLMM`, which has multiple methods (e.g., `print`, `summary`) associated with.
-Use `methods(class = 'communityPGLMM')` to see the full list.
-To visualize the design matrices' structures of random terms in the model object, we can use `communityPGLMM.plot.re()`.
-An example can be shown in Figure \@ref(fig:reStructure).
-It is also optional to simulate and plot data based on the design matrices with the argument `show.sim.image = TRUE`.
-
-(ref:reStructureCaption) A figure to show the structures of design matrices of random terms in `communityPGLMM`.
-
-```{r reStructure, fig.width=6, fig.height=6, fig.cap="(ref:reStructureCaption)"}
-communityPGLMM.plot.re(x = pglmm_example, show.image = TRUE, show.sim.image = FALSE)
-```
-
-# Example analysis
-
-Simulation of data; then apply the main functions.
-
-## Phylogenetic diversity
-
-```{r}
-tree = phyr::phylotree
-phyr::psv(comm_b, tree) # comm_b is a site by species matrix
-phyr::psd(comm_b, tree)
-x = phyr::pcd(comm_b, tree, verbose = FALSE)
-```
-
-## Species' traits
-
-## Community assembly
-
-```{r}
-pglmm_example
-```
-
-# Benchmark
-
-A table to compare the speed of functions in `phyr` with their original versions.
-
-```{r, echo=FALSE}
-# data prep for pez::communityPGLMM, not necessary for phyr::communityPGLMM
-dat = arrange(dat, site, sp)
-nspp = n_distinct(dat$sp)
-nsite = n_distinct(dat$site)
-
-dat$site = as.factor(dat$site)
-dat$sp = as.factor(dat$sp)
-
-tree = ape::drop.tip(phylotree, setdiff(phylotree$tip.label, unique(dat$sp)))
-Vphy <- ape::vcv(tree)
-Vphy <- Vphy/max(Vphy)
-Vphy <- Vphy/exp(determinant(Vphy)$modulus[1]/nspp)
-Vphy = Vphy[levels(dat$sp), levels(dat$sp)]
-
-# prepare random effects
-re.site <- list(1, site = dat$site, covar = diag(nsite))
-re.sp <- list(1, sp = dat$sp, covar = diag(nspp))
-re.sp.phy <- list(1, sp = dat$sp, covar = Vphy)
-# sp is nested within site
-re.nested.phy <- list(1, sp = dat$sp, covar = Vphy, site = dat$site)
-re.nested.rep <- list(1, sp = dat$sp, covar = solve(Vphy), site = dat$site) # equal to sp__@site
-# can be named 
-re = list(re.sp = re.sp, re.sp.phy = re.sp.phy, re.nested.phy = re.nested.phy, re.site = re.site)
-```
-
-```{r eval=T}
-microbenchmark::microbenchmark(phyr::pcd(comm = comm_a, tree = phylotree, reps = 1000, verbose = F),
-                               picante::pcd(comm = comm_a, tree = phylotree, reps = 1000),
-                               times = 30)
-rt = ape::rtree(n = 1000)
-microbenchmark::microbenchmark(phyr::vcv2(rt, corr = T),
-                               ape::vcv(rt, corr = T))
-
-microbenchmark::microbenchmark(phyr::psv(comm_a, phylotree, compute.var = T),
-                               picante::psv(comm_a, phylotree, compute.var = T))
-
-microbenchmark::microbenchmark(
-  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
-                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = T, 
-                       optimizer = "bobyqa"),
-    phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
-                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = T, 
-                       optimizer = "Nelder-Mead"),
-  phyr::communityPGLMM(pa ~ 1 + shade + (1|sp__) + (1|site) + (1|sp__@site), dat, 
-                       family = "binomial", cov_ranef = list(sp = phylotree), REML = F, cpp = F, 
-                       optimizer = "Nelder-Mead"),
-  pez::communityPGLMM(pa ~ 1 + shade, data = dat, family = "binomial", sp = dat$sp, 
-                      site = dat$site, random.effects = re, REML = F),
-  times = 10
+```{r eval=FALSE}
+pglmm(
+  Y ~ trait * env +
+    (1 | sp__) +
+    (1 | site__) +
+    (trait | site) +
+    (env | sp__) +
+    (1 | sp__@site),
+  data = data,
+  cov_randef = list(sp = phy.sp, site = V.space),
+  family = 'binomial',
+  bayes = FALSE,
+  REML = TRUE
 ) 
@@
 
-# Implications
+Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1.
+The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures.
+The independent variables `trait` and `env` take on different values for each species and site, respectively.
+Sites and species (`sp`) are treated as random effects: thus, `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites.
+For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species.
+The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺.
+A covariance matrix 𝚺 is specified by `cov_randef = list(sp = phy.sp, site = V.space)`.
+The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape`.
+It is also possible to specify an explicit covariance matrix, such as `site = V.space`.
+To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between two species is proportional to the distance between them on the phylogenetic tree.
 
-# Resources
+The syntax `(1|sp__)` or  `(1|site__)` generates two random effects, one without and one with phylogenetic or spatial covariances; in contrast, `(1|sp)` would generate only a single random effect that is independent among species.
+`pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in 𝚺 even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of 𝚺.
+Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence.
+To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable.
+Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix 𝚺.
+Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site.
+Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local, giving the covariances among species only within sites.
+Other forms of a nested term are available in `pglmm()`, which can be used to study more complicated questions such as bipartite networks.
 
-This package is available at Github.
+With `bayes = FALSE`, `pglmm()` is fit using a frequentist approach.
+ML or REML is used for fitting, with `REML = TRUE` as the default.
+For a non-Gaussian model (e.g., `family = 'binomial'`), an iterated quasi-likelihood method is used for model fitting which gives the approximate likelihood; p-values for the fixed effects are given by a Wald test and for the random effects by profile likelihood, although we recommend bootstrap-based tests when computationally feasible.
+Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer`) due to the algorithm used.
+With `bayes = TRUE`, a Bayesian approach is implemented using INLA [@rue2009approximate], which gives parameter estimates and credible intervals.
+For large problems with the number of species-site combinations exceeding 2000, the Bayesian computations are considerably faster than the frequentist computations.
+Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects.
+Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects.
 
-# Future directions
+## *`cor_phylo()`*
 
-In the future, we intend to ...
+`cor_phylo()` makes it possible to compare suites of traits among species, accounting for their phylogenetic relatedness.
+To identify suites of traits under joint selection, such as traits that together make up adaptive syndromes, it is necessary to perform a correlation analysis in which phylogenetic relatedness is factored out.
+`cor_phylo()` does this.
+It can also include within-species variation (e.g., measurement error) which should better-expose the underlying correlations in traits among species.
+Whereas `pglmm()` can be used to identify the composition of communities within a region, `cor_phylo()` can be used to assess patterns of traits among species that make up the regional species pool.
+
+The syntax for `cor_phylo()` is
+
+```{r eval=FALSE}
+cor_phylo(
+  variates = ~ trait1 + trait2,
+  species = ~ sp,
+  phy = phy.sp,
+  covariates = list(trait1 ~ env),
+  meas_errors = list(trait1 ~ me1, trait2 ~ me2),
+  data = data,
+  boot = 2000
+)
+```
+
+In this example, the correlation between two traits, `trait1` and `trait2`, is assessed, and the column named `sp` in `data` identifies the species.
+The object `phy.sp` specifies the phylogenetic covariance matrix as a 'phylo' object from the `ape` package.
+`cor_phylo()` estimates the phylogenetic signal for each trait by assuming that trait evolution is given by a Ornstein-Uhlenbeck process.
+The term `covariates = list(trait1 ~ env)` includes the independent variable `env` for `trait1`, to remove possible confounding effects; only an intercept is estimated if no covariate is provided for a trait.
+Within-species variation (measurement error) is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree.
+If within-species standard errors are not provided for a given trait, the trait values are assumed to be known without error.
+Finally, `cor_phylo()` can perform parametric bootstrapping to give confidence intervals for all parameter estimates: correlations, phylogenetic signals, covariate coefficients, and coefficient covariances.
+
+# Relationships to other methods and software
+
+`pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` and `ape`, respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility.
+Both have new syntax that makes them more intuitive to use.
+`pglmm()` also has new associated functions that plot the design of the covariance matrices, making model interpretation easier.
+Both are now coded in C++ (for key functions), which speeds computation time by 5-10X.
+`pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA that is particularly useful for large datasets.
+Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.
+
+## *`pglmm()`*
+
+`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4`, although it allows the specification of phylogenetic covariance matrices.
+`pglmm()` also allows "nested" models which arise when phylogenetic covariances only act within single communities, rather than among communities.
+Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present.
+Nested models are structurally incompatible with the architecture of `lme4`.
+
+There are alternative programs to `pglmm()`, although they have limitations that `pglmm()` overcomes.
+@hadfield2013tale use the R package `MCMCglmm` [@mcmcglmm2010] to perform phylogenetic community analyses, although they also use ASReml because its penalized quasi-likelihood (PQL) approach is computationally much faster.
+Hierarchical Modelling of Species Communities (HMSC-R) [@tikhonov2019joint] performs community analyses using Bayesian MCMC approaches, although it does not include nested terms.
+It is also possible to code up specific phylogenetic community models using flexible Bayesian platforms such as WinBugs, Stan, and JAGS, although this will involve considerable programming and expertise.
+
+### pglmm as a Joint Species Distribution Model (JSDM)
+
+pglmm is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny [@wilkinson2019comparison].
+
+## *`cor_phylo()`*
+
+The R package `mvMORPH` [@clavel2015mvmorph] can fit a broad range of models, of which `cor_phylo()` can be formulated as a special case.
+While `cor_phylo()` does not have the flexibility of `mvMORPH`, it is correspondingly simpler to use.
+Also, `cor_phylo()` has built-in bootstrapping capabilities that are necessary to give confidence in the parameter estimates and p-values.
+The function `evolvcv.lite()` in the R package `phytools` [@revell2012phytools] will compute phylogenetic correlations, and changes in phylogenetic correlations through time [see also @caetano2018estimating], although the phylogenetic covariance matrix is derived under the assumption of Brownian motion evolution.
+This contrasts `cor_phylo()` in which the strength of phylogenetic signal is computed at the same time as the correlation.
+It is also possible to code the `cor_phylo()` model using platforms such as WinBugs, Stan, and JAGS.
+
+# Example usage
+
+We simulated datasets to demonstrate how to use `pglmm()` and `cor_phylo()`.
+Details about simulations can be found in the Appendix.
+Our goal of this section is to provide some general ideas about the input and output of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results.
+For the late purposes, please see the package vignettes and @ives2018book.
+
+## _`pglmm()`_
+
+We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites (we focused on abundance but presence/absence can also be investigated by setting `family = 'binomial'`) while accounting for species’ phylogenetic relationships and site spatial autocorrelations (by setting `cov_ranef = list(species = phy, site = Vspace)`).
+This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.
+
+```{r}
+z <- pglmm(
+  abund ~ 1 + envi + trait + envi:trait +
+    (1 | species__)  + (1 | site__) + 
+    (envi | species__) + (trait | site) +
+    (1 | species__@site),
+  data = dat, 
+  cov_ranef = list(species = phy, site = Vspace)
+)
+summary(z)
+```
+
+The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other statistics based on it (e.g., AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties (Standard errors, Z scores, and P values if fitted by maximum likelihood or lower and upper boundaries of credible intervals).
+
+(ref:designPlotCap) The structures of design matrices of random terms in a phylogenetic generalized linear mixed model with `r nspp` species and `r nsite` sites.
+
+```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
+knitr::include_graphics(normalize_path("designPlot.pdf"))
+```
+
+## *`cor_phylo()`*
+
+Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) of 50 species.
+We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95 respectively.
+
+```{r echo=FALSE}
+source("simulation_cor_phylo.R")
+```
+
+```{r}
+z2 <- cor_phylo(variates = ~ trait_1 + trait_2, 
+                covariates = list(trait_2 ~ cov_trait_2),
+                species = ~ sp, phy = phy, 
+                meas_errors = list(trait_1 ~ se_trait_1, trait_2 ~ se_trait_2), 
+                data = traits)
+z2
+```
+
+# Future development
+
+Develop interface to fit model with Stan via brm?
+
+<!-- # Example analysis
+
+To demonstrate the usage of main functions in `phyr`, we simulated a dataset with 50 species and 30 communities.
+We first simulated a coalescent phylogeny of 50 species with function `ape::rcoal`.
+For each species, we then simulated one continuous functional trait along the phylogeny.
+We also simulated one environmental variable with all 30 communities located evenly along the gradient.
+Environmental variable, functional trait, and their interaction all determine the abundance of species over sites following the model below.
+
+$$y=(\beta_{0}+b_{0})+(\beta_{1}+b_{1})envi+(\beta_{2}+b_{2})trait+(\beta_{3}+b_{3})envi*trait+e$$
+$$b_{0}\sim Gaussian(0,\sigma_{0}^{2}\Sigma_{spp})$$
+$$b_{1}\sim Gaussian(0,\sigma_{1}^{2}\Sigma_{spp})$$
+$$b_{2}\sim Gaussian(0,\sigma_{2}^{2}\Sigma_{site})$$
+$$e\sim Gaussian(0,\sigma_{e}^2)$$
+
+Here, we set all coefficients (beta_0 to beta_3) to be 1; we also set variances of all random terms to be 1.
+$\Sigma_{spp}$ is a variance-covariance matrix converted from the phylogeny.
+$\Sigma_{site}$ is an identity matrix since we treat sites as independent replications.
+In the above model, different species have different overall abundance (intercept) but closely related species have similar overall abundance.
+Abundance of different species also change differently along the environmental gradient; however, closely related species again change similarly in response to environmental change.
+We also included a random term for functional trait; therefore the relationship between functional traits and species abundance varied across sites independently to account for unmeasured environmental variables.
+For demonstration purpose, we only simulated one functional trait and one environmental variable; in real datasets, multiple environmental variables and multiple functional traits can be included in the model.
+However, more random terms will come with higher computational burdens and the program will take longer to finish.
+-->
 

### 2019-09-19 simmulation_pglmm.R

@@ -1,225 +1,0 @@
-# devtools::install_github("daijiang/phyr")
-
-library(ape)
-library(LatticeKrig)
-
-#######################################################
-############## Set up simulation dataset ##############
-#######################################################
-
-###### Set up species and site variables
-nspp <- 20
-nsite <- 15
-
-# construct phylogeny from 'ape'
-phy <- rcoal(nspp)
-phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
-
-trait <- rTraitCont(phy, sigma = .1)
-
-# generate environmental site variable
-envi <- matrix(1:nsite, nrow = 1, ncol = nsite)
-envi <- (envi - mean(envi)) / sd(envi)
-
-# create spatial coordinates for sites
-# (to make a visually more clear spatial covariance matrix, sites are placed on a line)
-x.coord <- rep(0, nsite)
-y.coord <- sort(runif(n = nsite))
-
-# create data.frame with species, sites, trait, and environmental variable
-dat = data.frame(
-  species = rep(phy$tip.label, nsite),
-  site = rep(1:nsite, each = nspp),
-  x.coord = rep(x.coord, each = nspp),
-  y.coord = rep(y.coord, each = nspp),
-  envi = rep(envi, each = nspp),
-  trait = rep(trait, times = nsite)
-)
-
-###### Set up model parameters for simulations
-
-# parameters
-beta0 <- beta1 <- beta2 <- beta3 <- 1 # fixed terms
-sd.b0 <- sd.b1 <- sd.b2 <- sd.nested <- sd.resid <- 1 # sd of random terms
-
-# whether or not to include phylogenetic signal in B0 and B1
-signal.b0 <- TRUE
-signal.b1 <- TRUE
-
-# set up species-specific regression coefficients as random effects
-if (signal.b0) {
-  b0 <- beta0 + rTraitCont(phy, sigma = sd.b0 / 10)
-} else {
-  b0 <- beta0 + rnorm(nspp, sd = sd.b0)
-}
-if (signal.b1) {
-  b1 <- beta1 + rTraitCont(phy, sigma = sd.b1 / 10)
-} else {
-  b1 <- beta1 + rnorm(nspp, sd = sd.b1)
-}
-b2 <- beta2 + rnorm(nsite, sd = sd.b2)
-
-# set up data.frame for parameters
-B <- data.frame(
-  b0 = rep(b0, nsite),
-  b1 = rep(b1, nsite),
-  b2 = rep(b2, each = nspp),
-  beta3 = beta3
[…]

### 2019-09-19 simulation_cor_phylo.R

@@ -1,0 +1,52 @@
+library(ape)
+
+set.seed(123)
+# Set up parameter values for simulating data
+n <- 50
+phy <- rcoal(n, tip.label = 1:n)
+trt_names <- paste0("par", 1:2)
+
+R <- matrix(c(1, 0.7, 0.7, 1), nrow = 2, ncol = 2) # true var-corv matrix
+d <- c(0.3, 0.95) # true phylo signal
+B2 <- 1
+
+Se <- c(0.2, 1) # measurement error
+M <- matrix(Se, nrow = n, ncol = 2, byrow = TRUE)
+colnames(M) <- trt_names
+
+# Set up needed matrices for the simulations
+p <- length(d)
+
+star <- stree(n)
+star$edge.length <- array(1, dim = c(n, 1))
+star$tip.label <- phy$tip.label
+
+Vphy <- vcv(phy)
+Vphy <- Vphy/max(Vphy)
+Vphy <- Vphy/exp(determinant(Vphy)$modulus[1]/n)
+
+tau <- matrix(1, nrow = n, ncol = 1) %*% diag(Vphy) - Vphy
+C <- matrix(0, nrow = p * n, ncol = p * n)
+for (i in 1:p) for (j in 1:p) {
+  Cd <- (d[i]^tau * (d[j]^t(tau)) * (1 - (d[i] * d[j])^Vphy))/(1 - d[i] * d[j])
+  C[(n * (i - 1) + 1):(i * n), (n * (j - 1) + 1):(j * n)] <- R[i, j] * Cd
+}
+MM <- matrix(M^2, ncol = 1)
+V <- C + diag(as.numeric(MM))
+iD <- t(chol(V))
+
+XX <- iD %*% rnorm(2 * n)
+X <- matrix(XX, n, p)
+colnames(X) <- trt_names
+
+U <- list(cbind(rnorm(n, mean = 2, sd = 10)))
+names(U) <- trt_names[2]
+
+X[,2] <- X[,2] + B2[1] * U[[1]][,1] - B2[1] * mean(U[[1]][,1])
+
+traits = as.data.frame(X)
+names(traits) = c("trait_1", "trait_2")
+traits$sp = phy$tip.label
+traits$cov_trait_2 = U$par2[,1]
+traits$se_trait_1 = M[,1]
+traits$se_trait_2 = M[,2]

### 2019-09-19 simulation_pglmm.R

@@ -1,0 +1,174 @@
+# devtools::install_github("daijiang/phyr")
+
+library(ape)
+
+#######################################################
+############## Set up simulation dataset ##############
+#######################################################
+
+###### Set up species and site variables
+nspp <- 5
+nsite <- 10
+
+# construct phylogeny from 'ape'
+phy <- rcoal(nspp)
+phy <- compute.brlen(phy, method = "Grafen", power = 0.5)
+# set up phylogenetic attraction covariance matrix which requires a nested covariance matrix for phylogenetic attraction
+Vphy <- vcv(phy)
+Vphy <- Vphy / max(Vphy)
+Vphy <- Vphy / exp(determinant(Vphy)$modulus[1] / nspp)
+
+# Perform a Cholesky decomposition of Vphy. This is used to generate phylogenetic
+# signal: a vector of independent normal random variables, when multiplied by the
+# transpose of the Cholesky deposition of Vphy will have covariance matrix
+# equal to Vphy. Another way is to use mvtnorm package.
+cholVnested = t(chol(kronecker(diag(1, nrow = nsite), Vphy))) # nested component
+
+
+###### Set up covaraince matrices for simulations of spatial and nested covariances
+# create spatial coordinates for sites
+# (to make a visually more clear spatial covariance matrix, sites are placed on a line)
+x.coord <- rep(0, nsite)
+y.coord <- sort(runif(n = nsite))
+
+# Set up spatial correlation covariance matrix
+Dist <- matrix(0, nrow = nsite, ncol = nsite)
+for (i in 1:nsite)
+  for (j in 1:nsite)
+    Dist[i, j] <-
+  ((x.coord[i] - x.coord[j]) ^ 2 + (y.coord[i] - y.coord[j]) ^ 2) ^ .5
+
+range <- .25
+sd.space <- 1
+Vspace <- sd.space ^ 2 * exp(-Dist / range)
+rownames(Vspace) <- 1:nsite
+colnames(Vspace) <- 1:nsite
+
+cholVspace = t(chol(kronecker(Vspace, diag(1, nrow = nspp)))) # nested spatial correlation component
+
+# generate environmental site variable
+envi <- matrix(1:nsite, nrow = 1, ncol = nsite)
+envi <- (envi - mean(envi)) / sd(envi)
+
+trait <- rTraitCont(phy, sigma = .1)
+
[…]

### 2019-09-20 ms.Rmd
add closing marks

@@
 
-# Overview of 'phyr'
+# Overview of `phyr`
 
@@
 species phylogenetic covariance matrix nested within site covariance matrix).
-Finally, `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions.
+`pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions.
+Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g.
+via R package `brm`).
 
@@
   data = data,
-  cov_randef = list(sp = phy.sp, site = V.space),
+  cov_ranef = list(sp = phy.sp, site = V.space),
   family = 'binomial',
@@
 The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺.
-A covariance matrix 𝚺 is specified by `cov_randef = list(sp = phy.sp, site = V.space)`.
+A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`.
 The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape`.
@@
 Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects.
-Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects.
+Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects (Fig.
+\@ref(fig:designPlot)).
+
+(ref:designPlotCap) The structures of design matrices of random terms in a phylogenetic generalized linear mixed model with `r nspp` species and `r nsite` sites.
+
+```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
+knitr::include_graphics(normalize_path("designPlot.pdf"))
+```
 
@@
 Both have new syntax that makes them more intuitive to use.
-`pglmm()` also has new associated functions that plot the design of the covariance matrices, making model interpretation easier.
+`pglmm()` also has new associated functions that plot the design of the covariance matrices (Fig.
+\@ref(fig:designPlot)), making model interpretation easier.
 Both are now coded in C++ (for key functions), which speeds computation time by 5-10X.
@@
 This contrasts `cor_phylo()` in which the strength of phylogenetic signal is computed at the same time as the correlation.
-It is also possible to code the `cor_phylo()` model using platforms such as WinBugs, Stan, and JAGS.
+It is also possible to code the `cor_phylo()` model using platforms such as WinBugs, Stan, and JAGS; but again, this will require considerable programming and expertise.
 
@@
 Our goal of this section is to provide some general ideas about the input and output of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results.
-For the late purposes, please see the package vignettes and @ives2018book.
+For those purposes, please see the package vignettes and @ives2018book.
 
@@
 
-We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites (we focused on abundance but presence/absence can also be investigated by setting `family = 'binomial'`) while accounting for species’ phylogenetic relationships and site spatial autocorrelations (by setting `cov_ranef = list(species = phy, site = Vspace)`).
+We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites (we focused on abundance and used the default family of data distribution [Gaussian] but other distributions can also be specified by setting the `family` argument) while accounting for species’ phylogenetic relationships and site spatial autocorrelations (by setting `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic variance-covariance matrix, `Vspace` is a matrix of spatial distances among sites).
 This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.
@@
 
-(ref:designPlotCap) The structures of design matrices of random terms in a phylogenetic generalized linear mixed model with `r nspp` species and `r nsite` sites.
-
-```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
-knitr::include_graphics(normalize_path("designPlot.pdf"))
-```
-
 ## *`cor_phylo()`*
@@
 Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) of 50 species.
-We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95 respectively.
+We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.
 
@@
 
-# Future development
+The output of `cor_phylo()` includes log likelihood values, AIC, BIC, estimated correlation matrix of traits, estimated phylogenetic signals of traits, estimated coefficients and their uncertainties (standard errors, Z scores, and p values).
+If bootstrapping was enabled by setting the `boot` argument, the lower and upper boundaries of correlations, phylogenetic signal values, and coefficients will be appended.
 
-Develop interface to fit model with Stan via brm?
+# Closing remarks
+
+In recent years, there has been increasing effort to apply model-based approaches in community ecology.
+One particular example is the joint species distribution models.
+Despite the well-known importance of phylogenetic relationship in structuring species distributions and community composition, relatively few studies have incorporated phylogenetic relationships in model-based analyses of species distributions and community ecology.
+One potential reason is the lack of easy-to-use tools to facilitate the usage of phylogenetic species-distribution modelling in ecological communities.
+The package `phyr` fills this gap by providing implementations of phylogenetic species-distribution models with flexible model formula syntax (`pglmm()`).
+It also includes other model-based functions that are useful for ecological studies such as calculating community phylogenetic diversity (e.g.
+`psv()`) and estimating correlations among functional traits while accounting for their evolutionary history (`cor_phylo()`) (Table \@ref(tab:mainFunc)).
+
+The model formula of `pglmm()` is general and can be applied to other tools to fit phylogenetic species-distribution models.
+In the future, we plan to develop interfaces to fit such models with more bayesian programs such as Stan (based on the R package `brm`) [@brmsR2018].
+It is our hope that the `phyr` package and the proposed model formula for phylogenetic species-distribution models will facilitate the usage of such models by other researchers.
 
@@
 
+Funding for this work was provided by the National Science Foundation (NSF ...).
+
 # Authors' contributions
+
+D.L and A.R.I conceived the idea. All authors wrote the software and package documentations. All authors wrote the manuscript.
 
@@
 
+No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and will be submitted to CRAN in the near future.
+
 # References

### 2019-09-20 ms.Rmd
update closing remarks [skip ci]

@@
 
-The model formula of `pglmm()` is general and can be applied to other tools to fit phylogenetic species-distribution models.
-In the future, we plan to develop interfaces to fit such models with more bayesian programs such as Stan (based on the R package `brm`) [@brmsR2018].
-It is our hope that the `phyr` package and the proposed model formula for phylogenetic species-distribution models will facilitate the usage of such models by other researchers.
+The model formula of `pglmm()` is general and can be applied using other tools to fit phylogenetic species-distribution models.
+Thus, `pglmm()` can serve the developer community as a shell for new methods that fit GLMMs, with `phyr` providing an easy user interface.
+We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.
 

### 2019-09-20 ms.Rmd
update summary

@@
     css: style.css
-    keep_md: yes
+    keep_md: no
   bookdown::word_document2: 
@@
 1.
-Phylogenetic relationships among species are both a challenge and an opportunity for community ecology and comparative biology.
-Model-based phylogenetic analysis has advantages ...
-Currently, R functions to conduct these analyses were distributed across different packages and were mostly written in R, which can be very slow with large datasets.
+Model-based approaches are getting more and more popular in ecological studies, with joint species distribution models as a particular example.
+However, most of such studies did not include phylogenetic relationships among species despite its well-known importance in shaping species distribution and community composition.
+One potential reason is the lack of tools to allow ecologists to fit phylogenetic species distribution models easily.
 2.
-To help remedy this situation, we created `phyr`, an R package that collects and updates functions to conduct model-based phylogenetic analyses.
-By collecting these functions into one package and rewrite core functions in c++, this package can facilitate the use of model-based phylogenetic analyses.
-3. This paper presents the main functions of the package and provides an example using simulated data.
-4. The `phyr` package provides a unified software environment with improved performance to conduct model-based phylogenetic analyses in Ecology.
+To fill this gap, the R package `phyr` provides a general formula syntax and implementations of phylogenetic generalized linear mixed models.
+It also provides updated functions to calculate community phylogenetic diversity and correlations among functional traits while accounting for phylogenetic relationships.
+3.
+Other programs can also be used to fit phylogenetic species distribution models, though they require considerable programming and expertise.
+The model formula syntax we proposed here can be applied to those programs and thus serves as a unified framework for phylogenetic species distribution modelling.
+4. `phyr` can help current and future researchers formulate and analyze phylogenetic species-distribution models.
 

### 2019-09-24 simulation_pglmm.R
pdf to png [skip ci]

@@ -146,7 +146,7 @@ z <- pglmm(
 #   s2.init = sqrt(z$ss)
 # )
 # 
-pdf("designPlot.pdf", width = 10, height = 10)
+png("designPlot.png", width = 10, height = 10, units = "in", res = 300)
 pglmm.plot.re(
   x = z,
   sp.var = "species",

### 2019-09-27 ms.Rmd
pdf to png [skip ci]
Added my section on JSDMs

@@
 ```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
-knitr::include_graphics(normalize_path("designPlot.pdf"))
+knitr::include_graphics(normalize_path("designPlot.png"))
 ```
@@
 
-pglmm is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny [@wilkinson2019comparison].
+Joint Species Distribution Models (JSDMs) are models where the response is some distribution of multiple species across sites or samples (measured as abundances or occurrences), where all species are modelled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so].
+A recent review and evaluation of JSDMs, for example, only considers models that allow modelling residual dependencies among species [@wilkinson2019comparison].
+
+`pglmm` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny.
+JSDMs, and Species Distribution Models (SDM) in general, have typically been focused on producing accurate predictions of how species are distributed, usually in a geographic context.
+However, they can also be used for making inferences about the biology of species, such as which environmental factors are important in a species or set of species' biology, or whether traits or evolutionary history can help explain the structure of species communities.
+It is this kind of inference that is the focus of the JSDM implemented in `pglmm`.
+There is often a trade-off between improving predictions and making solid inferences, because increasing the complexity or flexibility of a model can improve its predictive power, but this same complexity makes it more difficult to understand what biology is being represented by the model outputs.
+In `pglmm`, the focus is on understanding whether evolutionary history is important in structuring species assemblages, which is facilitated by using a simply phylogenetic covariance model.
+This is in contrast to many other JSDMs [e.g.
+those in described in @wilkinson2019comparison], which attempt to flexibly estimate all pairwise species covariances, without taking into account any relationships amongst species, in an attempt to maximize predictive power.
+
+Of course, the goals of solid inference and prediction are not mutually exclusive.
+Good prediction requires avoiding overfitting, which can be facilitated by reducing the number of parameters in the model.
+In some systems, it may be possible to make better predictions using a simple phylogenetic model if phylogeny is a strong predictor of species covariance, or if many species are poorly sampled and thus estimating covariances between them individually results in higher prediction variance.
+Using phylogeny can help closely related species share statistical strength through phylogenetic partial pooling.
+Ultimately, it can be very powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals, both of which are highly important in modern ecology, and more easily and quickly draw mutual insights between them.
 

### 2019-09-27 ms.Rmd
update ms [skip ci]

@@
 
-Joint Species Distribution Models (JSDMs) are models where the response is some distribution of multiple species across sites or samples (measured as abundances or occurrences), where all species are modelled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so].
+Joint Species Distribution Models (JSDMs) are models where the response is distribution (abundances or occurrences) of multiple species across sites or samples, where all species are modelled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so].
 A recent review and evaluation of JSDMs, for example, only considers models that allow modelling residual dependencies among species [@wilkinson2019comparison].
 
-`pglmm` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny.
+`pglmm()` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny.
 JSDMs, and Species Distribution Models (SDM) in general, have typically been focused on producing accurate predictions of how species are distributed, usually in a geographic context.
 However, they can also be used for making inferences about the biology of species, such as which environmental factors are important in a species or set of species' biology, or whether traits or evolutionary history can help explain the structure of species communities.
-It is this kind of inference that is the focus of the JSDM implemented in `pglmm`.
+It is this kind of inference that is the focus of the JSDM implemented in `pglmm()`.
 There is often a trade-off between improving predictions and making solid inferences, because increasing the complexity or flexibility of a model can improve its predictive power, but this same complexity makes it more difficult to understand what biology is being represented by the model outputs.
-In `pglmm`, the focus is on understanding whether evolutionary history is important in structuring species assemblages, which is facilitated by using a simply phylogenetic covariance model.
+In `pglmm()`, the focus is on understanding whether evolutionary history is important in structuring species assemblages, which is facilitated by using a phylogenetic covariance model.
 This is in contrast to many other JSDMs [e.g.

### 2019-10-06 ms.Rmd
add citations [skip ci]

@@
 For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake.
-The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition.
+The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic].
 
@@
 For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured.
-Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain.
+Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@ives2006statistics].
 
@@
 
-Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities.
+Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized].
 The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities.
@@
 We then compare them to methods and programs that are currently available.
-Finally, we give applications of `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.
+Finally, we applied `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.
 
@@
 
-`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and phylogenetic diversity metrics (e.g.
+`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g.
 `psv()`, `pse()`).
 The workhorse functions of all groups are written in C++ to increase computational speed.
-Here, we will focus on the first two groups of functions because they are more complicated than phylogenetic diversity metrics.
+Here, we will focus on the first two groups of functions because they are more complicated than community phylogenetic diversity metrics.
 
@@
 `pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species.
-The syntax for `pglmm()` resembles that used in the R package `lme4`, and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`.
+The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`.
 `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be a phylogenetic covariance matrix or any other covariance matrix that the user defines (e.g., spatial or temporal autocorrelation matrix).
@@
 Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g.
-via R package `brm`).
+via R package `brm` [@brmsR2018]).
 
@@
 A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`.
-The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape`.
+The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018].
 It is also possible to specify an explicit covariance matrix, such as `site = V.space`.
@@
 Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local, giving the covariances among species only within sites.
+This nested term can be used to test for community clustering or overdispersion [@webb2002phylogenies; @ives2011generalized].
 Other forms of a nested term are available in `pglmm()`, which can be used to study more complicated questions such as bipartite networks.
 
-With `bayes = FALSE`, `pglmm()` is fit using a frequentist approach.
+With `bayes = FALSE`, `pglmm()` is fitted using a frequentist approach.
 ML or REML is used for fitting, with `REML = TRUE` as the default.
@@
 
-`pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` and `ape`, respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility.
+Earlier versions of `pglmm()` and `cor_phylo()` both appear in existing R packages (`pez` [@pearse2015pez] and `ape` [@ape2018], respectively), although the versions in `phyr` represent considerable improvements in ease-of-use, computational speed, and flexibility.
 Both have new syntax that makes them more intuitive to use.

### 2019-11-14 ms.Rmd
update ms [skip ci]

@@
 # opts_knit$set(root.dir = normalizePath("../"))
-opts_chunk$set(fig.align='center', cache = FALSE, warning = FALSE, 
+opts_chunk$set(fig.align='center', cache = TRUE, warning = FALSE, 
                message = FALSE, echo = TRUE)
@@
 1.
-Model-based approaches are getting more and more popular in ecological studies, with joint species distribution models as a particular example.
-However, most of such studies did not include phylogenetic relationships among species despite its well-known importance in shaping species distribution and community composition.
-One potential reason is the lack of tools to allow ecologists to fit phylogenetic species distribution models easily.
+Model-based approaches are increasingly popular in ecological studies.
+A good example of this trend is the use of joint species distribution models to ask questions about ecological communities.
+However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition.
+In part, this is due to lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
 2.
-To fill this gap, the R package `phyr` provides a general formula syntax and implementations of phylogenetic generalized linear mixed models.
-It also provides updated functions to calculate community phylogenetic diversity and correlations among functional traits while accounting for phylogenetic relationships.
+To fill this gap, the R package `phyr` provides implementations of phylogenetic generalized linear mixed models.
+It also provides functions to calculate community phylogenetic diversity, correlations among functional traits while accounting for phylogenetic relationships, and generalized models for analyzing phylogenetic comparative data.
 3.
-Other programs can also be used to fit phylogenetic species distribution models, though they require considerable programming and expertise.
-The model formula syntax we proposed here can be applied to those programs and thus serves as a unified framework for phylogenetic species distribution modelling.
-4. `phyr` can help current and future researchers formulate and analyze phylogenetic species-distribution models.
+Other programs can also be used to fit phylogenetic species distribution models, but they require considerable programming and expertise.
+The model formula syntax we propose serves as a simple and unified framework for phylogenetic species distribution modelling.
+4. `phyr` can help current and future researchers formulate and analyze phylogenetic joint species-distribution models.
 
-_Keywords_: phylogenetic diversity, phylogenetic generalized linear mixed models, functional trait, trait correlation, model-based methods
+_Keywords_: phylogenetic diversity, phylogenetic generalized linear mixed models, functional trait, trait correlation, model-based methods, Joint Species Distribution Model
 
@@
 These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep].
-For example, an insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common.
+For example, the larvae of an aquatic insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common.
 Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species.
 For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake.
-The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic].
+The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic; @losos1996; @webb2000; @webb2002phylogenies; @cavender2004; @helmus2007].
 
@@
 For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured.
-Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@ives2006statistics].
+Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@cressie1991; @ives2006statistics].
 
 How species respond to environmental factors, and how they respond to each other, depend on their traits.
-Therefore, relationships among functional traits can provide insights about the evolutionary history that has shaped species trait so they can occupy the same sites.
+Therefore, correlations among functional traits can provide insights about the evolutionary history that has shaped species traits so they can occupy the same sites.
 For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH.
 Is the positive correlation between these two traits caused by correlated selective forces?
-A challenge to answering this question is that phylogenetic correlations between trait values of closely related species can give the appearance of positive correlations between traits: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related.
-Therefore, it is important to account for species’ evolutionary history when studying correlations among functional traits.
+A challenge to answering this question is that phylogenetic correlations between trait values might reflect species phylogenetic relatedness rather than shared selection: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related.
+To distinguish between these two explanations -- convergence of suites of traits due to shared selective forces versus similarity due to phylogenetic relatedness -- it is necessary to account for phylogenies when performing correlation analyses between traits that could explain similarities in the distributions of species.
 
-Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized].
+Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized; @ovaskainen2011; @warton2015so].
 The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities.
@@
 These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow.
-The R package `phyr` is designed to overcome many of the costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance.
+The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance.
 Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`.
-We then compare them to methods and programs that are currently available.
-Finally, we applied `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.
+`pglmm()` allows the formulation of a diverses set of phylogenetic generalized linear models that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data.
+`cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait.
+We then compare `pglmm()` and `cor_phylo()` to methods and programs that are currently available.
+Finally, we apply  `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.
 
@@
 
-`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g.
-`psv()`, `pse()`).
+`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g., `psv()`, `pse()`).
 The workhorse functions of all groups are written in C++ to increase computational speed.
-Here, we will focus on the first two groups of functions because they are more complicated than community phylogenetic diversity metrics.
+Here, we will focus on the first two groups of functions, because they are more complicated and less readily available to practitioners than community phylogenetic diversity metrics.
 
@@
 The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`.
-`pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be a phylogenetic covariance matrix or any other covariance matrix that the user defines (e.g., spatial or temporal autocorrelation matrix).
-`pglmm()` can also fit models with “nested” covariance structures (e.g.
-species phylogenetic covariance matrix nested within site covariance matrix).
-`pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions.
+`pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be phylogenetic covariance matrices or any other covariance matrices that the user defines (e.g., spatial or temporal autocorrelation matrix).
+`pglmm()` can also fit models with “nested” covariance structures (e.g., a species phylogenetic covariance matrix nested within a site covariance matrix).
+`pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, binary, binomial or Poisson, and Bayesian mode with the addition of zero-inflated binomial and Poisson distributions.
 Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g.
-via R package `brm` [@brmsR2018]).
+via R package `brms` [@brmsR2018]).
 
@@
 The independent variables `trait` and `env` take on different values for each species and site, respectively.
-Sites and species (`sp`) are treated as random effects: thus, `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites.
+Sites and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites.
 For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species.
@@
 It is also possible to specify an explicit covariance matrix, such as `site = V.space`.
-To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between two species is proportional to the distance between them on the phylogenetic tree.
+To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between species is proportional to their shared evolutionary history (e.g., shared branch length on a phylogeny).
 
@@
 Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site.
-Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local, giving the covariances among species only within sites.
+Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local to sites, giving the covariances among species only within sites.
 This nested term can be used to test for community clustering or overdispersion [@webb2002phylogenies; @ives2011generalized].
@@
 ```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
-knitr::include_graphics(normalize_path("designPlot.png"))
+knitr::include_graphics(normalize_path("designPlot2.png"))
 ```
+
+Whereas `pglmm()` is designed to accept community composition data, in which the same species can occur in multiple sites, the algorithm used by `pglmm()` can equally be used for comparative data in which each species is represented by only a single data point.
+`pglmm.compare()` is a wrapper for `pglmm()` that is tailored for comparative data and thus provides an easy-to-use function for analyzing non-Gaussian phylogenetic data.
 
@@
 
-`cor_phylo()` makes it possible to compare suites of traits among species, accounting for their phylogenetic relatedness.
+`cor_phylo()` makes it possible to compare suites of traits among species, accounting for their phylogenetic relatedness [@zheng2009new; @johnson2014].
 To identify suites of traits under joint selection, such as traits that together make up adaptive syndromes, it is necessary to perform a correlation analysis in which phylogenetic relatedness is factored out.
@@
 
-In this example, the correlation between two traits, `trait1` and `trait2`, is assessed, and the column named `sp` in `data` identifies the species.
+In this example, the correlation between `trait1` and `trait2` is computed, and the column named `sp` in `data` identifies the species.
 The object `phy.sp` specifies the phylogenetic covariance matrix as a 'phylo' object from the `ape` package.
@@
 `pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4`, although it allows the specification of phylogenetic covariance matrices.
-`pglmm()` also allows "nested" models which arise when phylogenetic covariances only act within single communities, rather than among communities.
+`pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities.
 Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present.
@@
 Hierarchical Modelling of Species Communities (HMSC-R) [@tikhonov2019joint] performs community analyses using Bayesian MCMC approaches, although it does not include nested terms.
-It is also possible to code up specific phylogenetic community models using flexible Bayesian platforms such as WinBugs, Stan, and JAGS, although this will involve considerable programming and expertise.
+It is also possible to code specific phylogenetic community models using flexible Bayesian platforms such as WinBugs, Stan, and JAGS, although this will involve considerable programming and expertise.
 
@@
 
-Joint Species Distribution Models (JSDMs) are models where the response is distribution (abundances or occurrences) of multiple species across sites or samples, where all species are modelled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so].
-A recent review and evaluation of JSDMs, for example, only considers models that allow modelling residual dependencies among species [@wilkinson2019comparison].
-
-`pglmm()` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny.
+Joint Species Distribution Models (JSDMs) are models where the response variable is distribution (abundances or occurrences) of multiple species across sites or samples, where all species are modeled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so].
+`pglmm()` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modeled in a way that incorporates phylogenetic relatedness.
 JSDMs, and Species Distribution Models (SDM) in general, have typically been focused on producing accurate predictions of how species are distributed, usually in a geographic context.
-However, they can also be used for making inferences about the biology of species, such as which environmental factors are important in a species or set of species' biology, or whether traits or evolutionary history can help explain the structure of species communities.
+However, they can also be used for making inferences about the biology of species, such as which environmental factors are important in explaining the distribution of a species or set of species, and whether traits or evolutionary history can help explain these distributions.
 It is this kind of inference that is the focus of the JSDM implemented in `pglmm()`.
 There is often a trade-off between improving predictions and making solid inferences, because increasing the complexity or flexibility of a model can improve its predictive power, but this same complexity makes it more difficult to understand what biology is being represented by the model outputs.
-In `pglmm()`, the focus is on understanding whether evolutionary history is important in structuring species assemblages, which is facilitated by using a phylogenetic covariance model.
+By incorporating phylogenetic information, `pglmm()` has two uses.
+First, by identifying correlations that might be expected among species due to phylogeny, `pglmm()` gives better statistical properties for tests of factors underlying community composition.
+For example, @li2017statistical show that failure to account for phylogenetic correlations can inflate type I errors in tests for associating environmental variables and traits that underlie community composition.
+Second, `pglmm()` allows explicit focus on the importance of evolutionary history in structuring species assemblages, since phylogenetic covariances are explicitly estimated.
 This is in contrast to many other JSDMs [e.g.
-those in described in @wilkinson2019comparison], which attempt to flexibly estimate all pairwise species covariances, without taking into account any relationships amongst species, in an attempt to maximize predictive power.
+those in described in @wilkinson2019comparison], which attempt to estimate all pairwise species covariances without accounting for phylogeny.
 
@@
 Using phylogeny can help closely related species share statistical strength through phylogenetic partial pooling.
-Ultimately, it can be very powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals, both of which are highly important in modern ecology, and more easily and quickly draw mutual insights between them.
+Ultimately, it can be powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals, both of which are important in modern ecology, and more easily and quickly draw mutual insights between them.
 
@@
 Details about simulations can be found in the Appendix.
-Our goal of this section is to provide some general ideas about the input and output of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results.
+Our goal in this section is to provide some general ideas about the inputs and outputs of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results.
 For those purposes, please see the package vignettes and @ives2018book.
@@
 
-We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites (we focused on abundance and used the default family of data distribution [Gaussian] but other distributions can also be specified by setting the `family` argument) while accounting for species’ phylogenetic relationships and site spatial autocorrelations (by setting `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic variance-covariance matrix, `Vspace` is a matrix of spatial distances among sites).
+We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites.
+We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument.
+Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites.
 This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.
@@
 
-The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other statistics based on it (e.g., AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties (Standard errors, Z scores, and P values if fitted by maximum likelihood or lower and upper boundaries of credible intervals).
+The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other related statistics (AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties.
 
@@
 
-Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) of 50 species.
+Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) for 50 species.
 We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.
@@
 
-The output of `cor_phylo()` includes log likelihood values, AIC, BIC, estimated correlation matrix of traits, estimated phylogenetic signals of traits, estimated coefficients and their uncertainties (standard errors, Z scores, and p values).
+The output of `cor_phylo()` includes log-likelihood values, AIC, BIC, estimated correlation matrix of traits, estimated phylogenetic signals of traits, estimated coefficients and their uncertainties (standard errors, Z scores, and p values).
+In this example, the model gave good estimates of the parameters used to simulate the data.
 If bootstrapping was enabled by setting the `boot` argument, the lower and upper boundaries of correlations, phylogenetic signal values, and coefficients will be appended.
@@
 In recent years, there has been increasing effort to apply model-based approaches in community ecology.
-One particular example is the joint species distribution models.
-Despite the well-known importance of phylogenetic relationship in structuring species distributions and community composition, relatively few studies have incorporated phylogenetic relationships in model-based analyses of species distributions and community ecology.
-One potential reason is the lack of easy-to-use tools to facilitate the usage of phylogenetic species-distribution modelling in ecological communities.
+An example is the rise of joint species distribution models.
+Despite the well-known importance of phylogenetic relationships in structuring species distributions and community composition, relatively few studies have incorporated phylogenetic relationships in model-based analyses of species distributions and community ecology.
+A potential reason is the lack of easy-to-use tools to facilitate the use of phylogenetic species-distribution modeling in ecological communities.
 The package `phyr` fills this gap by providing implementations of phylogenetic species-distribution models with flexible model formula syntax (`pglmm()`).
-It also includes other model-based functions that are useful for ecological studies such as calculating community phylogenetic diversity (e.g.
-`psv()`) and estimating correlations among functional traits while accounting for their evolutionary history (`cor_phylo()`) (Table \@ref(tab:mainFunc)).
+It also includes other model-based functions that are useful for ecological studies such as estimating correlations among functional traits while accounting for their evolutionary history (`cor_phylo()`) and calculating community phylogenetic diversity (e.g.
+`psv()`) (Table \@ref(tab:mainFunc)).
 
@@
 Thus, `pglmm()` can serve the developer community as a shell for new methods that fit GLMMs, with `phyr` providing an easy user interface.
-We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.
+Using INLA as a backend to fit a Bayesian version of the pglmm model is an example of this approach.
+To facilitate this end, we are developing phyr openly on github and actively encourage community contribution.We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.
 
@@
 
-No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and will be submitted to CRAN in the near future.
+No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/web/packages/phyr/index.html).
 

### 2019-11-14 simulation_cor_phylo.R

@@ -1,6 +1,6 @@
 library(ape)
 
-set.seed(123)
+set.seed(1000)
 # Set up parameter values for simulating data
 n <- 50
 phy <- rcoal(n, tip.label = 1:n)

### 2019-11-14 simulation_pglmm.R

@@ -1,14 +1,14 @@
-# devtools::install_github("daijiang/phyr")
-
 library(ape)
+library(phyr)
 
 #######################################################
 ############## Set up simulation dataset ##############
 #######################################################
 
 ###### Set up species and site variables
-nspp <- 5
-nsite <- 10
+nspp <- 30
+nsite <- 30
+set.seed(1234)
 
 # construct phylogeny from 'ape'
 phy <- rcoal(nspp)
@@ -24,6 +24,9 @@ Vphy <- Vphy / exp(determinant(Vphy)$modulus[1] / nspp)
 # equal to Vphy. Another way is to use mvtnorm package.
 cholVnested = t(chol(kronecker(diag(1, nrow = nsite), Vphy))) # nested component
 
+trait <- rTraitCont(phy, sigma = .1)
+trait = rnorm(nspp)
+trait <- (trait - mean(trait)) / sd(trait)
 
 ###### Set up covaraince matrices for simulations of spatial and nested covariances
 # create spatial coordinates for sites
@@ -43,21 +46,20 @@ sd.space <- 1
 Vspace <- sd.space ^ 2 * exp(-Dist / range)
 rownames(Vspace) <- 1:nsite
 colnames(Vspace) <- 1:nsite
+Vspace <- Vspace / max(Vspace)
+Vspace <- Vspace / exp(determinant(Vspace)$modulus[1] / nsite)
+iV = t(chol(Vspace))
 
-cholVspace = t(chol(kronecker(Vspace, diag(1, nrow = nspp)))) # nested spatial correlation component
 
 # generate environmental site variable
 envi <- matrix(1:nsite, nrow = 1, ncol = nsite)
 envi <- (envi - mean(envi)) / sd(envi)
 
-trait <- rTraitCont(phy, sigma = .1)
 
 # create data.frame with species, sites, trait, and environmental variable
 dat = data.frame(
   species = rep(phy$tip.label, nsite),
   site = rep(1:nsite, each = nspp),
-  x.coord = rep(x.coord, each = nspp),
-  y.coord = rep(y.coord, each = nspp),
   envi = rep(envi, each = nspp),
   trait = rep(trait, times = nsite)
 )
@@ -66,8 +68,7 @@ dat = data.frame(
 
 # parameters
[…]

### 2019-11-24 ms.Rmd
update ms [skip ci]

@@
 title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
-author: "Daijiang Li, Russell Dinnage, Lucas Nell, Anthony Ives"
+author: "Daijiang Li, Russell Dinnage, Lucas Nell, Matthew Helmus, Anthony Ives"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
+  bookdown::word_document2: 
+    toc: no
   bookdown::html_document2:
@@
     keep_md: no
-  bookdown::word_document2: 
-    toc: no
   bookdown::tufte_html2:
@@
     toc: yes
+  redoc::redoc:
 fontsize: 12pt
@@
 2.
-To fill this gap, the R package `phyr` provides implementations of phylogenetic generalized linear mixed models.
-It also provides functions to calculate community phylogenetic diversity, correlations among functional traits while accounting for phylogenetic relationships, and generalized models for analyzing phylogenetic comparative data.
+To fill this gap, the R package `phyr` (pronounced fire) implements a suite of metrics, comparative methods and mixed models that use phylogenies to understand and predict community composition and other ecological and evolutionary phenomena.
+The `phyr` workhorse functions are implemented in C++ making all calculations and model estimations fast.
 3.
-Other programs can also be used to fit phylogenetic species distribution models, but they require considerable programming and expertise.
-The model formula syntax we propose serves as a simple and unified framework for phylogenetic species distribution modelling.
-4. `phyr` can help current and future researchers formulate and analyze phylogenetic joint species-distribution models.
+`phyr` can fit a variety of models such as phylogenetic joint-species distribution models, spatiotemporal-phylogenetic autocorrelation models, and phylogenetic trait‐based bipartite network models.
+`phyr` also estimates phylogenetically independent trait correlations with measurement error to test for adaptive syndromes and performs fast calculations of common alpha and beta phylogenetic diversity metrics.
+All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and the inclusion of phylogenetic terms based on phylogenetic covariance matrices.
+4. The functions and model formula syntax we propose in `phyr` serves as a simple and unified framework that ignites the use of phylogenies to address a variety of ecological questions.
 
@@
 Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`.
-`pglmm()` allows the formulation of a diverses set of phylogenetic generalized linear models that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data.
+`pglmm()` allows the formulation of a diverse set of phylogenetic generalized linear models (PGLMM) that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data.
 `cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait.
@@
 
-`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g., `psv()`, `pse()`).
+`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic comparative methods (`cor_phylo()` and `pglmm_compare()`), and community phylogenetic diversity metrics (e.g., `psv()`, `pse()`).
 The workhorse functions of all groups are written in C++ to increase computational speed.
-Here, we will focus on the first two groups of functions, because they are more complicated and less readily available to practitioners than community phylogenetic diversity metrics.
+Here, we will focus on the first two groups of functions (especially `pglmm()` and `cor_phylo()`), because they are more complicated and less readily available to practitioners than community phylogenetic diversity metrics.
 
@@
 read.csv("phyr_func.csv", check.names = FALSE) %>% 
-  knitr::kable(caption = "(ref:maunFuncCaption)", booktab = T) %>%
-  kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
-                latex_options = c("striped")) %>% 
-  kableExtra::column_spec(2, width = "25%")
+  knitr::kable(caption = "(ref:maunFuncCaption)") #%>%
+  # kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
+  #               latex_options = c("striped")) %>% 
+  # kableExtra::column_spec(2, width = "25%")
 ```
@@
 `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, binary, binomial or Poisson, and Bayesian mode with the addition of zero-inflated binomial and Poisson distributions.
-Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g.
-via R package `brms` [@brmsR2018]).
+Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan [e.g.
+via R package `brms` @brmsR2018].
 
@@
 Finally, a key to interpreting the results from a model is understanding the structure of the covariance matrices associated with the random effects.
-Therefore, `pglmm()` has associated plotting functions `pglmm.plot.ranef()` that present the design matrices for the random effects (Fig.
+Therefore, `pglmm()` has associated plotting functions `pglmm_plot_ranef()` that present the design matrices for the random effects (Fig.
 \@ref(fig:designPlot)).
@@
 
-```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
+```{r designPlot, echo = F, fig.cap="(ref:designPlotCap)", out.width="100%"}
 knitr::include_graphics(normalize_path("designPlot2.png"))
@@
 Whereas `pglmm()` is designed to accept community composition data, in which the same species can occur in multiple sites, the algorithm used by `pglmm()` can equally be used for comparative data in which each species is represented by only a single data point.
-`pglmm.compare()` is a wrapper for `pglmm()` that is tailored for comparative data and thus provides an easy-to-use function for analyzing non-Gaussian phylogenetic data.
+`pglmm_compare()` is a wrapper for `pglmm()` that is tailored for comparative data and thus provides an easy-to-use function for analyzing non-Gaussian phylogenetic data.
 
@@
 Both are now coded in C++ (for key functions), which speeds computation time by 5-10X.
-`pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA that is particularly useful for large datasets.
+`pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA [@rue2009approximate] that is particularly useful for large datasets.
 Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.
@@
 
-`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4`, although it allows the specification of phylogenetic covariance matrices.
+`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4` [@bates2015lme4], although it allows the specification of phylogenetic covariance matrices.
 `pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities.
@@
 
-### pglmm as a Joint Species Distribution Model (JSDM)
+### PGLMM as a Joint Species Distribution Model (JSDM)
 
@@
 Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites.
-This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.
+This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.
 
-```{r}
+```{r, eval=FALSE}
 z <- pglmm(
@@
 )
+```
+
+```{r, echo=FALSE}
+z = readRDS("z.rds")
+```
+
+```{r}
 summary(z)
@@
 The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other related statistics (AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties.
+These results show that `pglmm()` correctly recovered that the hypothetic functional trait interacted with environmental variable to affect species composition.
 
@@
 We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.
+We also set their measurement errors to be 0.2 and 1, respectively.
 
@@
 
-No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/web/packages/phyr/index.html).
+No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/package=phyr).
 

### 2019-11-24 simulation_pglmm.R

@@ -8,7 +8,7 @@ library(phyr)
 ###### Set up species and site variables
 nspp <- 30
 nsite <- 30
-set.seed(1234)
+set.seed(999)
 
 # construct phylogeny from 'ape'
 phy <- rcoal(nspp)
@@ -136,18 +136,18 @@ dat$abund <- y + rnorm(nspp * nsite, sd = sd.resid)
 # )
 # saveRDS(z.bayes, "z.bayes.rds")
 # 
-# system.time(z <- pglmm(
-#   abund ~ 1 + envi + trait + envi:trait +
-#     (1 | species__) + (1|site__) +
-#     (envi | species__) + 
-#     (trait | site) +
-#     (1 | species__@site),
-#   data = dat,
-#   cov_ranef = list(species = phy, site = Vspace),
-#   s2.init = c(0, 1, 0, 1, 0, 1, 1, 1),
-#   verbose = T
-# ))
-# saveRDS(z, "z.rds")
+system.time(z <- pglmm(
+  abund ~ 1 + envi + trait + envi:trait +
+    (1 | species__) + (1|site__) +
+    (envi | species__) +
+    (trait | site) +
+    (1 | species__@site),
+  data = dat,
+  cov_ranef = list(species = phy, site = Vspace),
+  s2.init = c(0, 1, 0, 1, 0, 1, 1, 1),
+  verbose = T
+))
+saveRDS(z, "z.rds")
 # 
 # png("designPlot2.png", width = 10, height = 10, units = "in", res = 300)
 # pglmm.plot.re(

### 2019-11-25 ms.Rmd
add appendix [skip ci]

@@
 
-<!-- # Example analysis
-
-To demonstrate the usage of main functions in `phyr`, we simulated a dataset with 50 species and 30 communities.
-We first simulated a coalescent phylogeny of 50 species with function `ape::rcoal`.
-For each species, we then simulated one continuous functional trait along the phylogeny.
-We also simulated one environmental variable with all 30 communities located evenly along the gradient.
-Environmental variable, functional trait, and their interaction all determine the abundance of species over sites following the model below.
-
-$$y=(\beta_{0}+b_{0})+(\beta_{1}+b_{1})envi+(\beta_{2}+b_{2})trait+(\beta_{3}+b_{3})envi*trait+e$$
-$$b_{0}\sim Gaussian(0,\sigma_{0}^{2}\Sigma_{spp})$$
-$$b_{1}\sim Gaussian(0,\sigma_{1}^{2}\Sigma_{spp})$$
-$$b_{2}\sim Gaussian(0,\sigma_{2}^{2}\Sigma_{site})$$
-$$e\sim Gaussian(0,\sigma_{e}^2)$$
-
-Here, we set all coefficients (beta_0 to beta_3) to be 1; we also set variances of all random terms to be 1.
-$\Sigma_{spp}$ is a variance-covariance matrix converted from the phylogeny.
-$\Sigma_{site}$ is an identity matrix since we treat sites as independent replications.
-In the above model, different species have different overall abundance (intercept) but closely related species have similar overall abundance.
-Abundance of different species also change differently along the environmental gradient; however, closely related species again change similarly in response to environmental change.
-We also included a random term for functional trait; therefore the relationship between functional traits and species abundance varied across sites independently to account for unmeasured environmental variables.
-For demonstration purpose, we only simulated one functional trait and one environmental variable; in real datasets, multiple environmental variables and multiple functional traits can be included in the model.
-However, more random terms will come with higher computational burdens and the program will take longer to finish.
--->
-
 # Acknowledgements
@@
 
+# Appendix
+
+To demonstrate the usage of main functions in phyr, we simulated a dataset with 30 species and 30 communities.
+We first simulated a coalescent phylogeny of 30 species with function `ape::rcoal()`.
+For each species, we then simulated one continuous functional trait along the phylogeny.
+We also simulated one environmental variable with all 30 communities located evenly along the gradient.
+Environmental variable, functional trait, and their interaction all determine the abundance of species over sites following the model below.
+
+$$y=(\beta_{0}+b_{0\_sp}+b_{0\_site})+(\beta_{1}+b_{1})envi+(\beta_{2}+b_{2})trait+\beta_{3}envi*trait+b_{3}+e$$
+$$b_{0\_{sp}}\sim Gaussian(0,\sigma_{0\_{sp}}^{2}\Sigma_{spp})$$
+$$b_{0\_{site}}\sim Gaussian(0,\sigma_{0\_{site}}^{2}\Sigma_{site})$$
+
+$$b_{1}\sim Gaussian(0,\sigma_{1}^{2}\Sigma_{spp})$$
+
+$$b_{2}\sim Gaussian(0,\sigma_{2}^{2}I_{site})$$
+$$b_{3}\sim Gaussian(0,\sigma_{3}^{2}\Sigma_{site}\otimes\Sigma_{spp})$$
+
+$$e\sim Gaussian(0,\sigma_{e}^2)$$
+
+Here, we set all coefficients ($\beta_0$ to $\beta_3$) to be 1; we also set variances of all random terms to be 1.
+$\Sigma_{spp}$ is a variance-covariance matrix converted from the phylogeny.
+$\Sigma_{site}$ is a vairance-covariance matrix among sites, which was converted from the site spatial distance matrix.
+$I_{site}$ is an identity matrix so that we treat sites as independent replications.
+$\Sigma_{site}\otimes\Sigma_{spp}$ a kronecker product and forces closely related species will be more likely to be observed in the same site (i.e.
+underdispersion).
+In the above model, different species have different overall abundance (intercept) but closely related species ($b_{0\_{sp}}$) and similar communities ($b_{0\_{site}}$) have similar overall abundance.
+Abundance of different species also change differently along the environmental gradient; however, closely related species again change similarly in response to environmental change.
+We included a random term for functional trait; therefore the relationship between functional traits and species abundance varied across sites independently to account for unmeasured environmental variables.
+For demonstration purposes, we only simulated one functional trait and one environmental variable; in real datasets, multiple environmental variables and multiple functional traits can be included in the model.
+However, more random terms will need more computational time to finish the model fitting.
+
 # References

### 2019-11-25 simulation_pglmm.R

@@ -136,18 +136,22 @@ dat$abund <- y + rnorm(nspp * nsite, sd = sd.resid)
 # )
 # saveRDS(z.bayes, "z.bayes.rds")
 # 
-system.time(z <- pglmm(
-  abund ~ 1 + envi + trait + envi:trait +
-    (1 | species__) + (1|site__) +
-    (envi | species__) +
-    (trait | site) +
-    (1 | species__@site),
-  data = dat,
-  cov_ranef = list(species = phy, site = Vspace),
-  s2.init = c(0, 1, 0, 1, 0, 1, 1, 1),
-  verbose = T
-))
-saveRDS(z, "z.rds")
+if(!file.exists("z.rds")){
+  z <- pglmm(
+    abund ~ 1 + envi + trait + envi:trait +
+      (1 | species__) + (1|site__) +
+      (envi | species__) +
+      (trait | site) +
+      (1 | species__@site),
+    data = dat,
+    cov_ranef = list(species = phy, site = Vspace),
+    s2.init = c(0, 1, 0, 1, 0, 1, 1, 1),
+    verbose = T
+  )
+  saveRDS(z, "z.rds")
+}
+
+
 # 
 # png("designPlot2.png", width = 10, height = 10, units = "in", res = 300)
 # pglmm.plot.re(

### 2020-02-13 ms.Rmd
update ms [skip ci]

@@
 title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
-author: "Daijiang Li, Russell Dinnage, Lucas Nell, Matthew Helmus, Anthony Ives"
+author: "Daijiang Li^1^, Russell Dinnage^2^, Lucas Nell^3^, Matthew Helmus^4^, Anthony Ives^3^"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
-  bookdown::word_document2: 
-    toc: no
   bookdown::html_document2:
@@
     keep_md: no
+  bookdown::word_document2: 
+    toc: no
+  bookdown::pdf_document2:
+    includes:
+      before_body: doc_prefix.tex
+      in_header: preamble.tex
+    keep_tex: yes
+    latex_engine: pdflatex
+    number_sections: no
+    toc: no
   bookdown::tufte_html2:
@@
 options(digits = 3, width = 88, knitr.kable.NA = '')
+tpl_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/template/preamble.tex'
+tpl_2 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/template/doc_prefix.tex'
+if(!file.exists(tpl_1f <- 'preamble.tex')) download.file(tpl_1, tpl_1f)
+if(!file.exists(tpl_2f <- 'doc_prefix.tex')) download.file(tpl_2, tpl_2f)
 # traits, phy, dat
@@
 Running Title: Model-based phylogenetic analyses
+
+^1^ Department of Wildlife Ecology & Conservation, University of Florida, Gainesville, FL 32611  
+^2^ Macroevolution & Macroecology Group, Research School of Biology, Australian National University, Canberra, ACT, Australia    
+^3^ Department of Integrative Biology, University of Wisconsin-Madison, Madison, WI 53706  
+^4^ Integrative Ecology Lab, Center for Biodiversity, Department of Biology, Temple University, Philadelphia, PA 19122
+
+Emails: daijianglee@gmail.com; r.dinnage@gmail.com; lucas.nell@wisc.edu; mrhelmus@temple.edu; arives@wisc.edu
 
@@
 `phyr` also estimates phylogenetically independent trait correlations with measurement error to test for adaptive syndromes and performs fast calculations of common alpha and beta phylogenetic diversity metrics.
-All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and the inclusion of phylogenetic terms based on phylogenetic covariance matrices.
+All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and phylogenetic terms are modelled as phylogenetic covariance matrices.
 4. The functions and model formula syntax we propose in `phyr` serves as a simple and unified framework that ignites the use of phylogenies to address a variety of ecological questions.
@@
 
-(ref:maunFuncCaption) List of main functions in `phyr` package.
+(ref:mainFuncCap) List of main functions in the `phyr` package.
 
@@
 read.csv("phyr_func.csv", check.names = FALSE) %>% 
-  knitr::kable(caption = "(ref:maunFuncCaption)") #%>%
-  # kableExtra::kable_styling(bootstrap_options = c("striped", "scale_down", "hover"),
-  #               latex_options = c("striped")) %>% 
-  # kableExtra::column_spec(2, width = "25%")
+  knitr::kable(caption = "(ref:mainFuncCap)", booktabs = T) %>%
+  kableExtra::kable_styling(latex_options = "striped", full_width = T) %>%
+  kableExtra::column_spec(3, width = "22em") 
 ```
@@
 
-`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species.
+Function `pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species.
 The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`.
@@
 Sites and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites.
-For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species.
-The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺.
-A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`.
+For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species (or some other correlation structure specified by the user).
+The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix $\Sigma$.
+A covariance matrix $\Sigma$ is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`.
 The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018].
-It is also possible to specify an explicit covariance matrix, such as `site = V.space`.
-To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between species is proportional to their shared evolutionary history (e.g., shared branch length on a phylogeny).
+To construct $\Sigma$ from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between species is proportional to their shared evolutionary history (e.g., shared branch length on a phylogeny).
+It is also possible to specify an explicit covariance matrix, such as `site = V.space`, where `V.space` is a covariance matrix created from the distance between sites.
 
 The syntax `(1|sp__)` or  `(1|site__)` generates two random effects, one without and one with phylogenetic or spatial covariances; in contrast, `(1|sp)` would generate only a single random effect that is independent among species.
-`pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in 𝚺 even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of 𝚺.
+`pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in $\Sigma$ even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of $\Sigma$.
 Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence.
 To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable.
-Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix 𝚺.
+Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix $\Sigma$.
 Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site.
@@
 For a non-Gaussian model (e.g., `family = 'binomial'`), an iterated quasi-likelihood method is used for model fitting which gives the approximate likelihood; p-values for the fixed effects are given by a Wald test and for the random effects by profile likelihood, although we recommend bootstrap-based tests when computationally feasible.
-Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer`) due to the algorithm used.
+Note that `REML = TRUE` is an option for non-Gaussian models (in contrast to `glmer()`) due to the algorithm used.
 With `bayes = TRUE`, a Bayesian approach is implemented using INLA [@rue2009approximate], which gives parameter estimates and credible intervals.
@@
 The term `covariates = list(trait1 ~ env)` includes the independent variable `env` for `trait1`, to remove possible confounding effects; only an intercept is estimated if no covariate is provided for a trait.
-Within-species variation (measurement error) is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree.
+Within-species variation is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree.
 If within-species standard errors are not provided for a given trait, the trait values are assumed to be known without error.
@@
 `pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4` [@bates2015lme4], although it allows the specification of phylogenetic covariance matrices.
-`pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities.
+`pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities; an example is illustrated by the `(1|sp__@site)` term in Fig.
+\@ref(fig:designPlot).
 Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present.
@@
 Using phylogeny can help closely related species share statistical strength through phylogenetic partial pooling.
-Ultimately, it can be powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals, both of which are important in modern ecology, and more easily and quickly draw mutual insights between them.
+Ultimately, it can be powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals and more easily and quickly draw mutual insights between them.
 
@@
 We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument.
-Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites.
+Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = V.space)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites.
 This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.
@@
 z <- pglmm(
-  abund ~ 1 + envi + trait + envi:trait +
+  abund ~ 1 + env + trait + env:trait +
     (1 | species__)  + (1 | site__) + 
-    (envi | species__) + (trait | site) +
+    (env | species__) + (trait | site) +
     (1 | species__@site),
   data = dat, 
-  cov_ranef = list(species = phy, site = Vspace)
+  cov_ranef = list(species = phy, site = V.space)
 )
@@
 Thus, `pglmm()` can serve the developer community as a shell for new methods that fit GLMMs, with `phyr` providing an easy user interface.
-Using INLA as a backend to fit a Bayesian version of the pglmm model is an example of this approach.
-To facilitate this end, we are developing phyr openly on github and actively encourage community contribution.We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.
+Using INLA as a backend to fit a Bayesian version of the PGLMM model is an example of this approach.
+To facilitate this end, we are developing phyr openly on github and actively encourage community contribution.
+We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.
 
@@
 
-Funding for this work was provided by the National Science Foundation (NSF ...).
+Funding for this work was provided by the National Science Foundation (US-NSF-DEB Dimensions of Biodiversity, 1240804).
 
@@
 
-No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/package=phyr).
+No data were used in this study.
+R code used to simulate data for PGLMMs as described in the Appendix is available at <https://github.com/daijiang/phyr_ms/blob/master/simulation_pglmm.R>.
+R code used to simulate data for comparative methods is available at <https://github.com/daijiang/phyr_ms/blob/master/simulation_cor_phylo.R>.
+`phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/package=phyr).
+
+\clearpage
 
@@
 
-To demonstrate the usage of main functions in phyr, we simulated a dataset with 30 species and 30 communities.
+To demonstrate the usage of main functions in `phyr`, we simulated a dataset with 30 species and 30 communities.
 We first simulated a coalescent phylogeny of 30 species with function `ape::rcoal()`.
@@
 We also simulated one environmental variable with all 30 communities located evenly along the gradient.
-Environmental variable, functional trait, and their interaction all determine the abundance of species over sites following the model below.
+The environmental variable, functional trait, and their interaction all determine the abundance of species among sites according to the model below:
 
@@
 $$b_{0\_{site}}\sim Gaussian(0,\sigma_{0\_{site}}^{2}\Sigma_{site})$$
-
 $$b_{1}\sim Gaussian(0,\sigma_{1}^{2}\Sigma_{spp})$$
-
 $$b_{2}\sim Gaussian(0,\sigma_{2}^{2}I_{site})$$
 $$b_{3}\sim Gaussian(0,\sigma_{3}^{2}\Sigma_{site}\otimes\Sigma_{spp})$$
-
 $$e\sim Gaussian(0,\sigma_{e}^2)$$
 
-Here, we set all coefficients ($\beta_0$ to $\beta_3$) to be 1; we also set variances of all random terms to be 1.
-$\Sigma_{spp}$ is a variance-covariance matrix converted from the phylogeny.
-$\Sigma_{site}$ is a vairance-covariance matrix among sites, which was converted from the site spatial distance matrix.
-$I_{site}$ is an identity matrix so that we treat sites as independent replications.
-$\Sigma_{site}\otimes\Sigma_{spp}$ a kronecker product and forces closely related species will be more likely to be observed in the same site (i.e.
+We set all coefficients ($\beta_0$ to $\beta_3$) to 1; we also set variances of all random terms to 1.
+$\Sigma_{spp}$ is a covariance matrix converted from the phylogeny.
+$\Sigma_{site}$ is a covariance matrix among sites, which was converted from the site spatial distance matrix.
+$I_{site}$ is the identity matrix so that we treat sites as independent replicates.
+$\Sigma_{site}\otimes\Sigma_{spp}$ is a matrix generated as a kronecker product, and it makes closely related species more likely to be observed in the same site (i.e.
 underdispersion).
-In the above model, different species have different overall abundance (intercept) but closely related species ($b_{0\_{sp}}$) and similar communities ($b_{0\_{site}}$) have similar overall abundance.
-Abundance of different species also change differently along the environmental gradient; however, closely related species again change similarly in response to environmental change.
-We included a random term for functional trait; therefore the relationship between functional traits and species abundance varied across sites independently to account for unmeasured environmental variables.
+Different species have different overall abundance (intercept) but closely related species ($b_{0\_{sp}}$) and similar communities ($b_{0\_{site}}$) have similar overall abundance.
+Abundances of different species also change differently along the environmental gradient; however, closely related species change similarly in response to the environmental variable.
+We included a random term for functional trait; therefore, the relationship between functional trait and species abundance varied across sites independently to account for unmeasured environmental variables.
 For demonstration purposes, we only simulated one functional trait and one environmental variable; in real datasets, multiple environmental variables and multiple functional traits can be included in the model.
-However, more random terms will need more computational time to finish the model fitting.
+
+\clearpage
 

### 2020-02-13 simulation_pglmm.R

@@ -43,24 +43,24 @@ for (i in 1:nsite)
 
 range <- .25
 sd.space <- 1
-Vspace <- sd.space ^ 2 * exp(-Dist / range)
-rownames(Vspace) <- 1:nsite
-colnames(Vspace) <- 1:nsite
-Vspace <- Vspace / max(Vspace)
-Vspace <- Vspace / exp(determinant(Vspace)$modulus[1] / nsite)
-iV = t(chol(Vspace))
+V.space <- sd.space ^ 2 * exp(-Dist / range)
+rownames(V.space) <- 1:nsite
+colnames(V.space) <- 1:nsite
+V.space <- V.space / max(V.space)
+V.space <- V.space / exp(determinant(V.space)$modulus[1] / nsite)
+iV = t(chol(V.space))
 
 
 # generate environmental site variable
-envi <- matrix(1:nsite, nrow = 1, ncol = nsite)
-envi <- (envi - mean(envi)) / sd(envi)
+env <- matrix(1:nsite, nrow = 1, ncol = nsite)
+env <- (env - mean(env)) / sd(env)
 
 
 # create data.frame with species, sites, trait, and environmental variable
 dat = data.frame(
   species = rep(phy$tip.label, nsite),
   site = rep(1:nsite, each = nspp),
-  envi = rep(envi, each = nspp),
+  env = rep(env, each = nspp),
   trait = rep(trait, times = nsite)
 )
 
@@ -124,13 +124,13 @@ dat$abund <- y + rnorm(nspp * nsite, sd = sd.resid)
 ############## Analyze simulation dataset #############
 #######################################################
 # z.bayes <- pglmm(
-#   abund ~ 1 + envi + trait + envi:trait +
+#   abund ~ 1 + env + trait + env:trait +
 #     (1 | species__)  + (1|site__) +
-#     (envi | species__) + 
+#     (env | species__) + 
 #     (trait | site) +
 #     (1 | species__@site),
 #   data = dat,
-#   cov_ranef = list(species = phy, site = Vspace),
+#   cov_ranef = list(species = phy, site = V.space),
 #   bayes = T,
 #   verbose = T
 # )
@@ -138,13 +138,13 @@ dat$abund <- y + rnorm(nspp * nsite, sd = sd.resid)
 # 
 if(!file.exists("z.rds")){
   z <- pglmm(
-    abund ~ 1 + envi + trait + envi:trait +
+    abund ~ 1 + env + trait + env:trait +
       (1 | species__) + (1|site__) +
[…]

### 2020-02-14 simulation_pglmm.R
sim pglmm [skip ci]

@@ -6,7 +6,7 @@ library(phyr)
 #######################################################
 
 ###### Set up species and site variables
-nspp <- 30
+nspp <- 60
 nsite <- 30
 set.seed(999)
 
@@ -42,7 +42,7 @@ for (i in 1:nsite)
   ((x.coord[i] - x.coord[j]) ^ 2 + (y.coord[i] - y.coord[j]) ^ 2) ^ .5
 
 range <- .25
-sd.space <- 1
+sd.space <- 1 # sd of b0_site
 V.space <- sd.space ^ 2 * exp(-Dist / range)
 rownames(V.space) <- 1:nsite
 colnames(V.space) <- 1:nsite
@@ -67,51 +67,56 @@ dat = data.frame(
 ###### Set up model parameters for simulations
 
 # parameters
-beta0 <- beta1 <- beta2 <- beta3 <- 1 # fixed terms
-sd.b0 <- sd.b1 <- sd.b2 <- sd.b3 <- sd.resid <- 1 # sd of random terms
+beta0 <- beta3 <- 1 # intercept and interaction terms
+beta1 <- beta2 <- 0.5 # main effects of envi and trait
+sd.b0_sp <- sd.resid <- 1 # sd of intercept random term and residual 
+sd.b1_sp <- 1 # sd of envi|sp__
+sd.b2_site <- 1 # sd of trait|site
+sd.b3_nested <- 1 # sd of sp__@site
+# sd of random terms
 
 # whether or not to include phylogenetic signal in B0 and B1
-signal.b0 <- TRUE
-signal.b1 <- TRUE
+signal.b0_sp <- TRUE
+signal.b1_sp <- TRUE
 
 # set up species-specific regression coefficients as random effects
 
-if (signal.b0) {
+if (signal.b0_sp) {
   # this accounts for the scaling of the variance terms by Vphy when det(Vphy) = 1
-  b0 <- rTraitCont(phy, sigma = sd.b0 * Vphy[1,1]^.5) 
-  b0 = b0 - mean(b0)
+  b0_sp <- rTraitCont(phy, sigma = sd.b0_sp * Vphy[1,1]^.5) 
+  b0_sp = b0_sp - mean(b0_sp)
 } else {
-  b0 <- rnorm(nspp, sd = sd.b0)
+  b0_sp <- rnorm(nspp, sd = sd.b0_sp)
 }
 
 b0_site = iV %*% rnorm(nsite, 0, 1)
 b0_site = b0_site - mean(b0_site)
 
[…]

### 2020-02-14 ms.Rmd
Aligning equations by ~ or =

@@
 
-$$y=(\beta_{0}+b_{0\_sp}+b_{0\_site})+(\beta_{1}+b_{1})envi+(\beta_{2}+b_{2})trait+\beta_{3}envi*trait+b_{3}+e$$
-$$b_{0\_{sp}}\sim Gaussian(0,\sigma_{0\_{sp}}^{2}\Sigma_{spp})$$
-$$b_{0\_{site}}\sim Gaussian(0,\sigma_{0\_{site}}^{2}\Sigma_{site})$$
-$$b_{1}\sim Gaussian(0,\sigma_{1}^{2}\Sigma_{spp})$$
-$$b_{2}\sim Gaussian(0,\sigma_{2}^{2}I_{site})$$
-$$b_{3}\sim Gaussian(0,\sigma_{3}^{2}\Sigma_{site}\otimes\Sigma_{spp})$$
-$$e\sim Gaussian(0,\sigma_{e}^2)$$
+\begin{align*}
+y = & ~ (\beta_{0}+b_{0\_sp}+b_{0\_site})+(\beta_{1}+b_{1})envi+(\beta_{2}+b_{2})trait+\beta_{3}envi*trait+b_{3}+e \\
+b_{0\_{sp}} \sim & ~ \text{Gaussian}(0,\sigma_{0\_{sp}}^{2}\Sigma_{spp}) \\
+b_{0\_{site}} \sim & ~ \text{Gaussian}(0,\sigma_{0\_{site}}^{2}\Sigma_{site}) \\
+b_{1} \sim & ~ \text{Gaussian}(0,\sigma_{1}^{2}\Sigma_{spp}) \\
+b_{2} \sim & ~ \text{Gaussian}(0,\sigma_{2}^{2}I_{site}) \\
+b_{3} \sim & ~ \text{Gaussian}(0,\sigma_{3}^{2}\Sigma_{site}\otimes\Sigma_{spp}) \\
+e \sim & ~ \text{Gaussian}(0,\sigma_{e}^2)
+\end{align*}
 

### 2020-02-23 ms.Rmd
Updated LAN email
submitted version [skip ci]

--- 2020-02-13 ms.Rmd
@@
 title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
-author: "Daijiang Li^1^, Russell Dinnage^2^, Lucas Nell^3^, Matthew Helmus^4^, Anthony Ives^3^"
+author: "Daijiang Li^1^, Russell Dinnage^2,3^, Lucas Nell^4^, Matthew R. Helmus^5^, Anthony Ives^4†^"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output:
+  bookdown::pdf_document2:
+    includes:
+      before_body: doc_prefix.tex
+      in_header: preamble.tex
+    keep_tex: yes
+    latex_engine: pdflatex
+    number_sections: no
+    toc: no
   bookdown::html_document2:
@@
   bookdown::word_document2: 
-    toc: no
-  bookdown::pdf_document2:
-    includes:
-      before_body: doc_prefix.tex
-      in_header: preamble.tex
-    keep_tex: yes
-    latex_engine: pdflatex
-    number_sections: no
     toc: no
@@
 
+\footnotesize
 ^1^ Department of Wildlife Ecology & Conservation, University of Florida, Gainesville, FL 32611  
-^2^ Macroevolution & Macroecology Group, Research School of Biology, Australian National University, Canberra, ACT, Australia    
-^3^ Department of Integrative Biology, University of Wisconsin-Madison, Madison, WI 53706  
-^4^ Integrative Ecology Lab, Center for Biodiversity, Department of Biology, Temple University, Philadelphia, PA 19122
+^2^ Research School of Biology, Australian National University, Acton ACT 2601, Australia    
+^3^ Institute for Applied Ecology, University of Canberra, Bruce ACT 2617, Australia  
+^4^ Department of Integrative Biology, University of Wisconsin-Madison, Madison, WI 53706  
+^5^ Integrative Ecology Lab, Center for Biodiversity, Department of Biology, Temple University, Philadelphia, PA 19122
 
-Emails: daijianglee@gmail.com; r.dinnage@gmail.com; lucas.nell@wisc.edu; mrhelmus@temple.edu; arives@wisc.edu
+Emails: daijianglee@gmail.com; r.dinnage@gmail.com; lucas@lucasnell.com; mrhelmus@temple.edu; arives@wisc.edu   
+^†^ Corresponding author
+\normalsize
 
@@
 These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow.
+
 The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance.
@@
 The independent variables `trait` and `env` take on different values for each species and site, respectively.
-Sites and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites.
+Sites (`site`) and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites.
 For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species (or some other correlation structure specified by the user).
@@
 `pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in $\Sigma$ even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of $\Sigma$.
-Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence.
+Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence from a community.
 To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable.
@@
 (ref:designPlotCap) The structures of design matrices of random terms in a phylogenetic generalized linear mixed model with `r nspp` species and `r nsite` sites.
+Random terms `1|sp` and `1|site` allow different species or sites to have different intercepts, respectively.
+Random terms `1|sp__` and `1|site` allow closely related species or sites with similar conditions to have similar intercepts, respectively.
+Random term `env|sp` allows different species to have different environment-abundance relationships independently.
+Random term `1|sp__@site` is a nested term and allows closely related species more likely to occur in the same site.
 
@@
 We simulated datasets to demonstrate how to use `pglmm()` and `cor_phylo()`.
-Details about simulations can be found in the Appendix.
+Details about simulations of PGLMM can be found in the Appendix.
 Our goal in this section is to provide some general ideas about the inputs and outputs of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results.
@@
 We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument.
-Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = V.space)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites.
+Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a covriance matrix among sites.
 This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.
@@
   abund ~ 1 + env + trait + env:trait +
-    (1 | species__)  + (1 | site__) + 
-    (env | species__) + (trait | site) +
-    (1 | species__@site),
+    (1 | sp__)  + (1 | site__) + 
+    (env | sp) + (1 | sp__@site),
   data = dat, 
-  cov_ranef = list(species = phy, site = V.space)
+  cov_ranef = list(sp = phy, site = V.space)
 )
@@
 In recent years, there has been increasing effort to apply model-based approaches in community ecology.
-An example is the rise of joint species distribution models.
 Despite the well-known importance of phylogenetic relationships in structuring species distributions and community composition, relatively few studies have incorporated phylogenetic relationships in model-based analyses of species distributions and community ecology.
@@
 
-\clearpage
+<!-- \clearpage -->
 
@@
 
-To demonstrate the usage of main functions in `phyr`, we simulated a dataset with 30 species and 30 communities.
-We first simulated a coalescent phylogeny of 30 species with function `ape::rcoal()`.
+To demonstrate the usage of main functions in `phyr`, we simulated a dataset with `r nspp` species and `r nsite` communities.
+We first simulated a coalescent phylogeny of `r nspp` species with function `ape::rcoal()`.
 For each species, we then simulated one continuous functional trait along the phylogeny.
-We also simulated one environmental variable with all 30 communities located evenly along the gradient.
+We also simulated one environmental variable with all `r nsite` communities located evenly along the gradient.
 The environmental variable, functional trait, and their interaction all determine the abundance of species among sites according to the model below:
 
-$$y=(\beta_{0}+b_{0\_sp}+b_{0\_site})+(\beta_{1}+b_{1})envi+(\beta_{2}+b_{2})trait+\beta_{3}envi*trait+b_{3}+e$$
-$$b_{0\_{sp}}\sim Gaussian(0,\sigma_{0\_{sp}}^{2}\Sigma_{spp})$$
-$$b_{0\_{site}}\sim Gaussian(0,\sigma_{0\_{site}}^{2}\Sigma_{site})$$
-$$b_{1}\sim Gaussian(0,\sigma_{1}^{2}\Sigma_{spp})$$
-$$b_{2}\sim Gaussian(0,\sigma_{2}^{2}I_{site})$$
-$$b_{3}\sim Gaussian(0,\sigma_{3}^{2}\Sigma_{site}\otimes\Sigma_{spp})$$
-$$e\sim Gaussian(0,\sigma_{e}^2)$$
+\begin{align*}
+y = & ~ (\beta_{0}+b_{0\_sp}+b_{0\_site})+(\beta_{1}+b_{1})envi+\beta_{2}trait+\beta_{3}envi*trait+b_{2}+e \\
+b_{0\_{sp}} \sim & ~ \text{Gaussian}(0,\sigma_{0\_{sp}}^{2}\Sigma_{spp}) \\
+b_{0\_{site}} \sim & ~ \text{Gaussian}(0,\sigma_{0\_{site}}^{2}\Sigma_{site}) \\
+b_{1} \sim & ~ \text{Gaussian}(0,\sigma_{1}^{2}I_{spp}) \\
+b_{2} \sim & ~ \text{Gaussian}(0,\sigma_{3}^{2}\Sigma_{site}\otimes\Sigma_{spp}) \\
+e \sim & ~ \text{Gaussian}(0,\sigma_{e}^2)
+\end{align*}
 
@@
 $\Sigma_{site}$ is a covariance matrix among sites, which was converted from the site spatial distance matrix.
-$I_{site}$ is the identity matrix so that we treat sites as independent replicates.
+$I_{spp}$ is the identity matrix so that we treat species as independent replicates.
 $\Sigma_{site}\otimes\Sigma_{spp}$ is a matrix generated as a kronecker product, and it makes closely related species more likely to be observed in the same site (i.e.
@@
 Different species have different overall abundance (intercept) but closely related species ($b_{0\_{sp}}$) and similar communities ($b_{0\_{site}}$) have similar overall abundance.
-Abundances of different species also change differently along the environmental gradient; however, closely related species change similarly in response to the environmental variable.
-We included a random term for functional trait; therefore, the relationship between functional trait and species abundance varied across sites independently to account for unmeasured environmental variables.
+Abundances of different species also change differently along the environmental gradient.
+
 For demonstration purposes, we only simulated one functional trait and one environmental variable; in real datasets, multiple environmental variables and multiple functional traits can be included in the model.
 
-\clearpage
+<!-- We can also make closely related species change similarly in response to the environmental variable.
+Furthermore, we can include a random term for functional trait to allow the relationship between functional trait and species abundance vary across sites independently to account for unmeasured environmental variables.
+-->
+
+<!-- \clearpage -->
 

### 2020-02-23 simulation_pglmm.R

@@ -6,9 +6,9 @@ library(phyr)
 #######################################################
 
 ###### Set up species and site variables
-nspp <- 60
-nsite <- 30
-set.seed(999)
+nspp <- 30
+nsite <- 20
+set.seed(9999)
 
 # construct phylogeny from 'ape'
 phy <- rcoal(nspp)
@@ -32,7 +32,7 @@ trait <- (trait - mean(trait)) / sd(trait)
 # create spatial coordinates for sites
 # (to make a visually more clear spatial covariance matrix, sites are placed on a line)
 x.coord <- rep(0, nsite)
-y.coord <- sort(runif(n = nsite))
+y.coord <- runif(n = nsite)
 
 # Set up spatial correlation covariance matrix
 Dist <- matrix(0, nrow = nsite, ncol = nsite)
@@ -58,7 +58,7 @@ env <- (env - mean(env)) / sd(env)
 
 # create data.frame with species, sites, trait, and environmental variable
 dat = data.frame(
-  species = rep(phy$tip.label, nsite),
+  sp = rep(phy$tip.label, nsite),
   site = rep(1:nsite, each = nspp),
   env = rep(env, each = nspp),
   trait = rep(trait, times = nsite)
@@ -68,7 +68,7 @@ dat = data.frame(
 
 # parameters
 beta0 <- beta3 <- 1 # intercept and interaction terms
-beta1 <- beta2 <- 0.5 # main effects of envi and trait
+beta1 <- beta2 <- 1 # main effects of envi and trait
 sd.b0_sp <- sd.resid <- 1 # sd of intercept random term and residual 
 sd.b1_sp <- 1 # sd of envi|sp__
 sd.b2_site <- 1 # sd of trait|site
@@ -77,7 +77,7 @@ sd.b3_nested <- 1 # sd of sp__@site
 
 # whether or not to include phylogenetic signal in B0 and B1
 signal.b0_sp <- TRUE
-signal.b1_sp <- TRUE
+signal.b1_sp <- FALSE
 
 # set up species-specific regression coefficients as random effects
 
@@ -114,7 +114,8 @@ B <- data.frame(
 
 y <- beta0 + B$b0_sp + B$b0_site +
   (beta1 + B$b1_sp) * dat$env +
-  (beta2 + B$b2_site) * dat$trait +
+  (beta2) * dat$trait +
+  # (beta2 + B$b2_site) * dat$trait +
   beta3 * dat$env * dat$trait +
[…]

### 2020-07-13 ms.Rmd
minor edits on ms

@@
 title: "`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities"
-author: "Daijiang Li^1^, Russell Dinnage^2,3^, Lucas Nell^4^, Matthew R. Helmus^5^, Anthony Ives^4†^"
+author: "Daijiang Li^1^, Russell Dinnage^2,3^, Lucas Nell^4^, Matthew R. Helmus^5^, Anthony R. Ives^4†^"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
 if(!require("phyr")) devtools::install_github("daijiang/phyr")
-xfun::pkg_attach2(c("phyr", "knitr", "kableExtra", "kableExtra", "tidyverse"))
+xfun::pkg_attach2(c("phyr", "knitr", "tidyverse"))
 # opts_knit$set(root.dir = normalizePath("../"))
@@
                message = FALSE, echo = TRUE)
-options(digits = 3, width = 88, knitr.kable.NA = '')
+options(digits = 3, width = 88, knitr.kable.NA = '', 
+        knitr.graphics.auto_pdf = TRUE)
 tpl_1 = 'https://raw.githubusercontent.com/daijiang/workflow_demo/master/template/preamble.tex'
@@
 if(!file.exists(tpl_2f <- 'doc_prefix.tex')) download.file(tpl_2, tpl_2f)
+if(knitr::is_latex_output() | knitr::is_html_output()){
+  library(kableExtra)
+} else {
+  options(kableExtra.auto_format = FALSE) # for docx
+}
+
 # traits, phy, dat
@@
 However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition.
-In part, this is due to lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
+In part, this is due to a lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
 2.
@@
 We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument.
-Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a covriance matrix among sites.
+Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a covariance matrix among sites.
 This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.

### 2020-07-13 ms.Rmd
Minor edits to cor_phylo portion of ms

@@
 The term `covariates = list(trait1 ~ env)` includes the independent variable `env` for `trait1`, to remove possible confounding effects; only an intercept is estimated if no covariate is provided for a trait.
+Covariates are linear terms fit using ML or REML with the rest of the model.
 Within-species variation is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree.
@@
 Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) for 50 species.
-We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.
-We also set their measurement errors to be 0.2 and 1, respectively.
+We set the true correlation between these two traits to be 0.7 and their phylogenetic signals (via an Ornstein-Uhlenbeck process) to be 0.3 and 0.95, respectively.
+We also set their measurement errors to be 0.2 and 1, respectively, and assigned the covariate `cov_trait_2` to `trait_2` with a slope of 1.
 

### 2020-07-15 ms.Rmd
new intro

@@
 Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community.
+For example, the larvae of an aquatic insect species might only occur in a particular lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common.
 These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep].
-For example, the larvae of an aquatic insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common.
-Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species.
-For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake.
-The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic; @losos1996; @webb2000; @webb2002phylogenies; @cavender2004; @helmus2007].
+Differences in community composition among locations depends on the variation in abiotic and biotic characteristics of the locations [@whittaker1956vegetation].
+Continuing with our example of aquatic insects, an insect species might occur in two nearby lakes because they both have low pH, or because they have the same predator community.
+Nearby lakes might also have the same insect species if the adult insects readily disperse between lakes.
+Thus, understanding the composition of a community, and the variation in composition among communities, requires understanding both species traits and environmental variables that together govern what species can occur where, and whether they can get there.
 
-Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity.
-The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species.
-Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured.
-For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured.
-Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@cressie1991; @ives2006statistics].
-
-How species respond to environmental factors, and how they respond to each other, depend on their traits.
-Therefore, correlations among functional traits can provide insights about the evolutionary history that has shaped species traits so they can occupy the same sites.
-For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH.
-Is the positive correlation between these two traits caused by correlated selective forces?
-A challenge to answering this question is that phylogenetic correlations between trait values might reflect species phylogenetic relatedness rather than shared selection: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related.
-To distinguish between these two explanations -- convergence of suites of traits due to shared selective forces versus similarity due to phylogenetic relatedness -- it is necessary to account for phylogenies when performing correlation analyses between traits that could explain similarities in the distributions of species.
+The importance of species traits, environmental characteristics, and geographical location for community composition imply that statistical analyses of community composition should be capable of incorporating all of these types of factors.
+For any statistical analysis, a model specifies variables that might explain the presence or abundance of species (predictor or independent variables) and the statistical structure of the unexplained variation.
+Even though species traits may ultimately dictate whether a species can occur in a specific community, it is unlikely that all relevant traits are known and measured for every species.
+Consequently, the effects of these traits will be treated as "unexplained variance" in the model.
+Nonetheless, because phylogenetically related species often have similar trait values, the unexplained variance will likely have phylogenetic signal [@felsenstein1985phylogenies; @freckleton2002phylogenetic; @blomberg2003testing].
+Therefore, statistical models of community composition should account for phylogenies [@li2017can; @frishkoff2017phylogenetic; @losos1996; @webb2000; @webb2002phylogenies; @cavender2004; @helmus2007].
+Similarly, environmental factors differing among locations that are unknown or unmeasured might nonetheless be more similar among locations that are close to each other.
+This will generate spatial autocorrelation in the unexplained variance among communities [@cressie1991; @ives2006statistics].
+Spatial autocorrelation can also be generated by the dispersal of individuals among nearby communities [@moran1953statistical; @royama2012analytical].
+Thus, the analysis of community composition requires statistical models that can explicitly include species traits and environmental factors, and also include them implicitly in the form of phylogenetic relationships among species and geographical distances among locations in the unexplained variance terms of the model.
 
@@
 They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities.
-These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow.
 
-The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance.
+Analyzing the composition of communities in a phylogenetic context may generate additional statistical questions about the evolutionary processes that have generated the distribution of values of traits among species.
+For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH.
+Is the positive correlation between these two traits caused by correlated selective forces?
+A challenge to answering this type of question is that phylogenetic correlations between trait values might reflect species phylogenetic relatedness rather than shared selection: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related.
+To distinguish between these two explanations -- convergence of suites of traits due to shared selective forces versus similarity due to phylogenetic relatedness -- it is necessary to account for phylogenies when performing correlation analyses between traits that could explain similarities in the distributions of species.
+
+The R package `phyr` is designed to allow users to easily develop a rich collection of models for the analyses of ecological communities that include phylogenetic correlations among species and/or spatial correlation among locations.
 Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`.

### 2020-07-19 ms.Rmd
Added citation for cor_phylo covariate method.
update ms and ref

@@
 All `phyr` methods are united under Brownian motion or Ornstein-Uhlenbeck models of evolution and phylogenetic terms are modelled as phylogenetic covariance matrices.
-4. The functions and model formula syntax we propose in `phyr` serves as a simple and unified framework that ignites the use of phylogenies to address a variety of ecological questions.
+4. The functions and model formula syntax we propose in `phyr` provide an easy-to-use collection of tools that we hope will ignite the use of phylogenies to address a variety of ecological questions.
 
@@
 It is also possible to specify an explicit covariance matrix, such as `site = V.space`, where `V.space` is a covariance matrix created from the distance between sites.
+For example, if we assume that spatial correlations follow a Gaussian function, then the correlation in residuals from sites _i_ and _j_ located a distance $d_{ij}$ from each other is $\exp(-(d_{ij}/r)^2)$, where _r_ is the "range" giving how quickly spatial correlation decreases with distance [@besag1975estimation].
 
@@
 `pglmm()` now supports several non-Gaussian distributions and allows Bayesian analyses using INLA [@rue2009approximate] that is particularly useful for large datasets.
-Finally, both include more output; for example, both now include facilities to perform likelihood ratio tests and compute AIC and BIC values for model comparisons.
+Finally, both include more output; for example, both now report the log-likelihood and related measures AIC and BIC.
+The log-likelihood makes it possible to compare full models to reduced models with some parameters removed, using likelihood ratio test [e.g., @neter1989applied] to give the significance of a collection of parameters.
+AIC and BIC can be used for model selection [@burnham2002model].
 
@@
 Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present.
+This is the case in the distribution of fish among lakes: after accounting for the effect of pH excluding some species from some lakes, a pattern of exclusion occurs in which phylogenetically related species are less likely to occur in the same lakes [@helmus2010communities].
 Nested models are structurally incompatible with the architecture of `lme4`.
@@
 We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument.
-Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a covariance matrix among sites.
+Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(sp = phy, site = V.space)` where `sp` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `V.space` is a covariance matrix among sites.
 This model can also be fitted with Bayesian framework by setting `bayes = TRUE`, which is recommended when data set is large.

### 2020-07-20 ms.Rmd
revision

@@
 The term `covariates = list(trait1 ~ env)` includes the independent variable `env` for `trait1`, to remove possible confounding effects; only an intercept is estimated if no covariate is provided for a trait.
-Covariates are linear terms fit using ML or REML with the rest of the model.
+Covariates are linear terms fit using the generalized least squares estimation equation [@martins1997phylogenies].
 Within-species variation is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree.
@@
 `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/package=phyr).
+Its documentations are available at <https://daijiang.github.io/phyr/index.html>.
 
@@
 For demonstration purposes, we only simulated one functional trait and one environmental variable; in real datasets, multiple environmental variables and multiple functional traits can be included in the model.
+For examples using `pglmm()` to analyse empirical dataset, see the package vignette (<https://daijiang.github.io/phyr/articles/phyr_example_empirical.html>).
 
